---
title: "University of Idaho Statistics 519 Spring 2015 Homework #2"
author: "Author: David Erich Seamon - erichs@uidaho.edu "
date: "Due Date: 02/20/2015"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{enumitem}
- \usepackage{tikz,amsmath}
- \newcommand\hlight[1]{\tikz[overlay, remember picture,baseline=-\the\dimexpr\fontdimen22\textfont2\relax]\node[rectangle,fill=blue!50,rounded corners,fill opacity = 0.2,draw,thick,text opacity =1] {$#1$};} 
- \pagestyle{fancy}
- \usepackage{amsmath}
- \newcommand\sufr[3][0pt]{$\rule{0pt}{\dimexpr#1+1.4ex\relax}^\frac{#2}{#3}$}
- \fancyhead[CO,CE]{Erich Seamon - University of Idaho}
- \fancyfoot[CO,CE]{Statistics 519 - Homework 2 - 02/20/2015}
- \renewcommand{\headrulewidth}{0.4pt}% Default \headrulewidth is 0.4pt
- \renewcommand{\footrulewidth}{0.4pt}% Default \footrulewidth is 0pt
- \fancyfoot[LE,RO]{\thepage}
output:
  pdf_document:
    keep_tex: yes
    
    

---

\noindent {\bf ASSIGNED PROBLEMS: 4.1, 4.2, 4.4, 4.12, 4.13, 4.15, 4.17, 5.3, 5.12, 5.16 }

\begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}
    \item [Problem 4.1]
    
    \end{enumerate}
    \begin{enumerate}
    

    \item[] Consider the two covariance matrices
      

  
        \[ \Sigma_{1} = \begin{pmatrix} 14 & 8 & 3 \\
                               8 & 5 & 2 \\
                               3 & 2 & 1 \end{pmatrix},
          \Sigma_{2} = \begin{pmatrix} 6 & 6 & 1\\ 
    		     6 & 8 & 2 \\
 				      1 & 2 & 1 \end{pmatrix} \]
               
               Show that 
               
               \[ |\Sigma_{1}| < |\Sigma_{2}|\]
               
               and that
               
               \[tr(\Sigma_{2} < tr(\Sigma+{1}))\]
               
               Thus the generalized variance of population 2 is greater than the generalized variance population 1, even though the total variance is less. Comment of why this is true in terms of the variances and correlations.
               
     \item[] 
               Solution:
               
                \end{enumerate}
   
```{r}
Sigma1 <- matrix(c(14,8,3,8,5,2,3,2,1), ncol=3)
Sigma2 <- matrix(c(6,6,1,6,8,2,1,2,1), ncol=3)
det(Sigma1)
det(Sigma2)
sum(diag(Sigma1))
sum(diag(Sigma2))
```
   
   \begin{enumerate}
   
   \item[] Comment:  
   
   The generalized population variance, or \[|\Sigma|\] is a measure of the concentration of y's close to the mean of the matrix - or its multicollinearity.  The smaller the Sigma, the more concentrated y values are around the mean.  The trace of a variance matrix provides the sum of the total variances.  As such, an increase in the variance will result in a larger Sigma.  This explains why the matrix Sigma1 - has a lower Sigma than mattrix Sigma2, but a  higher variance.

    \end{enumerate}

   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 4.2]
    \end{enumerate}
    \begin{enumerate}

        \item[] For \[z = (T')^-1 (y - \mu)\] in (4.4), show that \[E(z) = 0\;and\;cov(z) = I\]
        
        Solution:
        
        For a standardized vector, z, T' is derived from the Cholesky decomposition of the variance-covariance matrix - Sigma = T'T.  In addition, T' may be found by substituting it for the symmetric square root matrix of the variance-covariance matrix.  In either case, it follows that:
        
        \[\Sigma(z) = E(E^{.5})^{-1} (y - \mu)\]
        \[= (\Sigma^{.5})^{-1} E(y - \mu)\]
        \[= (E^{.5})^{-1} (E(y) - \mu)\]
        \[= 0\]
        
        \[If\;y\;is\;N_{p}(\mu, \Sigma),\;then\;z\;is\;N_{p}(0,I)\]
        
 \end{enumerate}

   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 4.4]
    \end{enumerate}
    \begin{enumerate}

        \item[] For \[(y-\mu)'\Sigma^{-1}(y-\mu)\] in (4.3) and (4.6), show that \[E[(y-\mu)'\Sigma^{-1}(y-\mu)] = p\]
        
        \[\;\;Assume\;E(y) = \mu\;and\;cov(y) = \Sigma\]
        
        Solution: The Mahalanobis Distance is a scalar value - and as such, the above equation can be reduced to:
        
        \[=\; E[tr(y - \mu)' \Sigma^{-1}(y - \mu)]\]
        
        which translates to
        
        \[=\;tr(I_{p})\;=\;p\]
        
         \end{enumerate}

   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
  \item[]
  \item[]
                
  \item [Problem 4.12]
    \end{enumerate}
    \begin{enumerate}
        
   \item[] Suppose
  
  \[ y\;is\;N_{4}(\mu, \Sigma),\;where\]
    
    \[ \mu = \begin{pmatrix} -2  \\
                               3  \\
                               -1  \\
                               5 \end{pmatrix},
          \Sigma = \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & -6 \\
 				      3 & -3 & 2 & 3\\
               9 & -6 & 3 & 9 \end{pmatrix} \]
               
    \setlength{\itemindent}{.5in}

    \item[(a)] Find the distribution of:
    
    \[z = 4y_{1} - 2y_{2} + y_{3} - 3y_{4}\ \]
   \item[]
   Solution:
   
   \[ if\;y\;is\;N_{p}(\mu, \Sigma),\;then\;a'y\;is\;N(a' \mu, a'\Sigma a)\]
   
   \[a' = 4, -2, 1, -3\]
   
   \[a'\mu\;is\;\begin{pmatrix} 4 & -2 & 1 & -3 \end{pmatrix}
          \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & -6 \\
   			      3 & -3 & 2 & 3\\
               9 & -6 & 3 & 9 \end{pmatrix} \]
               
  \[a'\Sigma a\;is\;\begin{pmatrix} 4 \\ -2 \\ 1 \\ -3 \end{pmatrix}
  \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & -6 \\
     		      3 & -3 & 2 & 3\\
               9 & -6 & 3 & 9 \end{pmatrix}
               \begin{pmatrix} 4 & -2 & 1 & -3 \end{pmatrix}\]
               
               \[ a'y\;is\;N(-30, 153)\]
  
\end{enumerate}

```{r}
mean <- matrix(c(-2, 3, -1, 5), ncol=1)
sigma <- matrix(c(11, -8, 3, 9, -8, 9, -3, -6, 3, -3, 2, 3, 9, -6, 3, 9), ncol=4)
a <- matrix(c(4, -2, 1, -3), ncol=1)
at <- t(a)
at %*% mean #--multiplying a' by the mean vector
at %*% sigma %*% a #---multiplying a' by sigma and by a
```
 
\begin{enumerate}
              
    \setlength{\itemindent}{.5in}
    \item[]

    \item[(b)] Find the joint distribution of:
    
    \[z_{1} = 3y_{1} + y_{2} + y_{3} + y_{4}\;and\;z_{2} = -2y_{1} + 3y_{2} + y_{3} -2y_{4}\]
    
     Solution:
     
      \[ if\;y\;is\;N_{p}(\mu, \Sigma),\;then\;Ay\;is\;N(A \mu, A \Sigma A')\]
   
   \[A = \begin{pmatrix} 1 & 1 & 1 & 1 \\ -2 & 3 & 1 & -2 \end{pmatrix} \]
   
   \[A \mu = \begin{pmatrix} 1 & 1 & 1 & 1 \\ -2 & 3 & 1 & -2 \end{pmatrix} \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & 6 \\
       	      3 & -3 & 2 & 3\\
               9 & 6 & 3 & 9 \end{pmatrix}  \]
   
   \[A \mu = \begin{pmatrix} 5  \\ 2  \end{pmatrix} \]
               
  \[A\Sigma A'\;is\;\begin{pmatrix} 1 & 1 & 1 & 1 \\ -2 & 3 & 1 & -2 \end{pmatrix}
  \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & -6 \\
   			      3 & -3 & 2 & 3\\
               9 & -6 & 3 & 9 \end{pmatrix}
               \begin{pmatrix} 1 & -2 \\ 1 & 3 \\ 1 & 1 \\ 1 & -2 \end{pmatrix}\]
               
               \[ A \Sigma A' = \begin{pmatrix} 27 & -79 \\ -79 & 361 \end{pmatrix}\]
               
               \[Ay\;is\;N(\begin{pmatrix} 5  \\ 2  \end{pmatrix}, \begin{pmatrix} 27 & -79 \\ -79 & 361 \end{pmatrix})\]
     
       \end{enumerate}

```{r}
mean <- matrix(c(-2, 3, -1, 5), ncol=1)  #---mean
sigma <- matrix(c(11, -8, 3, 9, -8, 9, -3, -6, 3, -3, 2, 3, 9, -6, 
3, 9), ncol=4) #---cov matrix
A <- matrix(c(1, -2, 1, 3, 1, 1, 1, -2), ncol=4) #---creating z1 and z2 matrix
At <- t(A) #---transposing A
A %*% mean #---multiplying A' by the mean vector
A %*% sigma %*% At #---multiplying A by sigma and by A'
```
 
 \begin{enumerate}
     
\setlength{\itemindent}{.5in}
    \item[]

    \item[(c)] Find the joint distribution of:
    
    \[z_{1} = 3y_{1} + y_{2} - 4y_{3} - y_{4}\;and \]
    
    \[z_{2} = -y_{1} - 3y_{2} + y_{3} - 2y_{4}\;and \]
    
    \[z_{3} = 2y_{1} + 2y_{2} + 4y_{3} - 5y_{4}\;and \]
    
    Solution:
    
    \[ if\;y\;is\;N_{p}(\mu, \Sigma),\;then\;Ay\;is\;N(A \mu, A \Sigma A')\]
   
   \[A = \begin{pmatrix} 3 & 1 & -4 & -1 \\ -1 & -3 & 1 & -2 \\ 2 & 2 &4 & -5 \end{pmatrix} \]
   
   \[A \mu = \begin{pmatrix} 3 & 1 & -4 & -1 \\ -1 & -3 & 1 & -2 \\ 2 & 2 &4 & -5 \end{pmatrix} \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & 6 \\
               3 & -3 & 2 & 3\\
               9 & 6 & 3 & 9 \end{pmatrix}  \]
   
   \[A \mu = \begin{pmatrix} -4  \\ -18 \\-27  \end{pmatrix} \]
               
  \[A\Sigma A'\;is\;\begin{pmatrix} 3 & 1 & -4 & -1 \\ -1 & -3 & 1 & -2 \\ 2 & 2 &4 & -5 \end{pmatrix}
  \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & -6 \\
   			      3 & -3 & 2 & 3\\
               9 & -6 & 3 & 9 \end{pmatrix}
               \begin{pmatrix} 1 & -2 \\ 1 & 3 \\ 1 & 1 \\ 1 & -2 \end{pmatrix}\]
               
               \[ A \Sigma A' = \begin{pmatrix} 35 & -18 & -6 \\ -18 & 46 & 14 \\ -6 & 14 & 93 \end{pmatrix}\]
               
              \[Ay\;is\;N(\begin{pmatrix} -4  \\ -18 \\ -27  \end{pmatrix}, \begin{pmatrix} 35 & -18 & -6 \\ -18 & 46 & 14 \\ -6 & 14 & 93 \end{pmatrix})\]

\end{enumerate}

```{r}
mean <- matrix(c(-2, 3, -1, 5), ncol=1)  #---mean
sigma <- matrix(c(11, -8, 3, 9, -8, 9, -3, -6, 3, -3, 2, 3, 9, -6, 3, 
9), ncol=4) #---cov matrix
A <- matrix(c(3, -1, 2, 1, -3, 2, -4, 1, 4, -1, -2, 
-5), ncol=4) #---creating z1 and z2 matrix
At <- t(A) #---transposing A
A %*% mean #---multiplying A' by the mean vector
A %*% sigma %*% At #---multiplying A by sigma and by A'
```
 
 \begin{enumerate}
               
               
  \setlength{\itemindent}{.5in}
    \item[]

    \item[(d)] What is the distribution of \[y_{3}?\]
    
    Solution: By taking the mean vector value of y3 and the similar diagonal value of y3 from the aforementioned covariance matrix, you can determine the distribution of y3:
    
    \[y\;is\;N(-1, 2)\]
               
  
  \setlength{\itemindent}{.5in}
  
    \item[]

    \item[(e)] What is the joint distribution of \[y_{2}\;and\;y_{4}?\]
    
    Solution: By taking the mean vector values of y2 and y4 and the similar diagonal value of y2 and y4 from the aforementioned covariance matrix, you can determine the distribution of y2 and y4:
    
    \[y\;is\;N\begin{pmatrix} 3 \\
                               5  \end{pmatrix}, \begin{pmatrix} 9 & -6 \\
                               -6 & 9  \end{pmatrix}\]
                               
  \setlength{\itemindent}{.5in}
  
    \item[]

    \item[(f)] What is the joint distribution of \[y_{1},\; \frac{1}{4}(y_{1} + y_{2}), \frac{1}{3}(y_{1} + y_{2} + y_{3}),\;and\;\frac{1}{4}(y_{1} + y_{2} + y_{3} + y_{4}) \]
    
    Solution:
    
    \[ A = \begin{pmatrix} y_{1} \\
                               \frac{1}{2}y_{1} + \frac{1}{2}y_{2} \\ 
                               \frac{1}{3}y_{1} + \frac{1}{3}y_{2} + \frac{1}{3}y_{3} \\
                               \frac{1}{4}y_{1} + \frac{1}{4}y_{2} + \frac{1}{4}y_{3} + \frac{1}{4}y_{4} \end{pmatrix}\]
                               
      \[ A = \begin{pmatrix}   1 & 0 & 0 & 0 \\ 
                               \frac{1}{2} & \frac{1}{2} & 0 & 0 \\
                               \frac{1}{3} & \frac{1}{3} & \frac{1}{3} & 0 \\
                               \frac{1}{4} & \frac{1}{4} & \frac{1}{4} & \frac{1}{4} \end{pmatrix}
                               \]            
                               
      \[ A = \begin{pmatrix}   1 & 0 & 0 & 0 \\ 
                               \frac{1}{2} & \frac{1}{2} & 0 & 0 \\
                               \frac{1}{3} & \frac{1}{3} & \frac{1}{3} & 0 \\
                               \frac{1}{4} & \frac{1}{4} & \frac{1}{4} & \frac{1}{4} \end{pmatrix}
                               \begin{pmatrix} -2 \\
                               3 \\
                               -1 \\
                               5 \end{pmatrix}\;=\; 
                               \begin{pmatrix} -2 \\
                               .5 \\
                               0 \\
                               1.25 \end{pmatrix}\]                           
    

\end{enumerate}

```{r}
mean <- matrix(c(-2, 3, -1, 5), ncol=1)  #---mean
sigma <- matrix(c(11, -8, 3, 9, -8, 9, -3, -6, 3, -3, 2, 3, 9, 
-6, 3, 9), ncol=4) #---cov matrix
A <- matrix(c(1, 1/2, 1/3, 1/4, 0, 1/2, 1/3, 1/4, 0, 0, 1/3, 
1/4, 0, 0, 0, 1/4), ncol=4) #---transposing A
roundA <- round(A, digits=3)
At <- t(roundA)
round(A %*% mean, digits=3) #---multiplying A' by the mean vector
```
 
 \begin{enumerate}
 
   \setlength{\itemindent}{.5in}
  
    \item[]
 
       \[ A \Sigma A' = \begin{pmatrix}   1 & 0 & 0 & 0 \\ 
                               \frac{1}{2} & \frac{1}{2} & 0 & 0 \\
                               \frac{1}{3} & \frac{1}{3} & \frac{1}{3} & 0 \\
                               \frac{1}{4} & \frac{1}{4} & \frac{1}{4} & \frac{1}{4} \end{pmatrix}
                              \begin{pmatrix} 11 & -8 & 3 & 9 \\ 
             -8 & 9 & -3 & -6 \\
     		      3 & -3 & 2 & 3\\
               9 & -6 & 3 & 9 \end{pmatrix}
               \begin{pmatrix}   1 & \frac{1}{2} & \frac{1}{3} & \frac{1}{4} \\ 
                               0 & \frac{1}{2} & \frac{1}{3} & \frac{1}{4} \\
                               0 & 0 & \frac{1}{3} & \frac{1}{4}  \\
                               0 & 0 & 0 & \frac{1}{4}  \end{pmatrix}\]
                               \[ A \Sigma A' = \begin{pmatrix} 11 & 1.5 & 2 & 3.75 \\ 1.5 & 1 & .67 & .875 \\ 2 & .67 & .67 & 1 \\ 3.75 & .875 & 1 & 1.688 \end{pmatrix}\]     
                               
                              \[Ay\;is\;N_{4}(\begin{pmatrix} -2  \\ .5 \\ 0 \\1.25  \end{pmatrix}, \begin{pmatrix} 11 & 1.5 & 2 & 3.75 \\ 1.5 & 1 & .67 & .875 \\ 2 & .67 & .67 & 1 \\ 3.75 & .875 & 1 & 1.688 \end{pmatrix})\]

 
 \end{enumerate}

```{r}                    
round(A %*% sigma %*% At, digits=3) #---multiplying A by sigma and by A'           
```                 
  \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 4.13]
    \end{enumerate}
    \begin{enumerate}
        
   \item[] Suppose:
   
   \[y\;is\;N_{4}(\mu, \Sigma),\;with\;\mu\;and\;\Sigma\;given\;in\;the\;previous\;problem\]
   
     \setlength{\itemindent}{.5in}
  
    \item[]

    \item[(a)] Find a vector z such that:
    
    \[z = (T')^{-1}(y - \mu) is N_{4}(0,I)\]
    
    Solution: 
    
    \[If\;y\;is\;N_{p}(\mu, \Sigma),\;then\;z\;is\;N_{p}(0,I)\]
    
    Using the cholesky decomposition, we can determine T from Sigma.  Putting T into our equation
    
    \[z = (T')^{-1}(y - \mu) is N_{4}(0,I)\]
    
    we generate a cholesky decomposed T matrix, and use that, in concert with the means of y1-y4, to determine z.
    
   \[ z = \begin{pmatrix}   .302 & 0 & 0 & 0 \\ 
                               .408 & .561 & 0 & 0 \\
                               -.87 & .261 & 1.015 & 0  \\
                               -.858 & -.343 & -.686 & .972  \end{pmatrix}
                              \begin{pmatrix} y + 2 \\ 
             y - 3 \\
       	      y + 1\\
               y - 5 \end{pmatrix}\]

    
        \end{enumerate}

    
```{r}
mean <- matrix(c(-2, 3, -1, 5), ncol=1)  #---mean
sigma <- matrix(c(11, -8, 3, 9, -8, 9, -3, -6, 3, -3, 2, 3, 9, -6, 
3, 9), ncol=4) #---cov matrix
T <- chol(sigma)
Tt <- t(T)
inverse_Tt <- solve(Tt)
inverse_Tt
```

 \begin{enumerate}

     \setlength{\itemindent}{.5in}
  
    \item[]

    \item[(b)] Find a vector z such that:
    
    \[z = (\Sigma^\frac{1}{2})^{-1} (y - \mu)\;is\;N_{4}(0,I)\]
    
    Solution:  Since A is positive definite, spectral decomposition was used, in concert with the means of y1-y4, to determine z.
    
    \[ \Sigma^\frac{1}{2} = CD^\frac{1}{2} C'\]
            \end{enumerate}
            
            
\[ z = (y - \mu)\;=\; \begin{pmatrix}   .810 & .305 & .143 & -.479 \\ 
                               .305 & .582 & .249 & 1.083 \\
                               .143 & .249 & 1.153 & -.298  \\
                               -.480 & -.083 & -.298 & .787  \end{pmatrix}
                              \begin{pmatrix} y + 2 \\ 
             y - 3 \\
               y + 1\\
               y - 5 \end{pmatrix}\]

```{r}
  ei = eigen(sigma)
  d = ei$values
      d = (d+abs(d))/2
      d2 = 1/sqrt(d)
      d2[d == 0] = 0
      (ei$vectors %*% diag(d2) %*% t(ei$vectors))
```
     
 \begin{enumerate}
\setlength{\itemindent}{.5in}
  
    \item[]

    \item[(c)] What is the distribution of:
    
    \[(y - \mu)' \Sigma^{-1} (y - \mu)?\]
    
    Solution:  The Mahalanobis Distance is known to have a chi-square distibution.
    
    \[If\;y\;is\;N_{p}(\mu, \Sigma),\;then\;(y - \mu)' \Sigma^{-1} (y - \mu)\;is\;\chi_{p}^{2}\]
    
    \[(y - \mu)' \Sigma^{-1} (y - \mu) = (y - \mu)' \Sigma^\frac{1}{2}\Sigma^\frac{1}{2} (y - \mu) = \chi_{p}^{2}\]
    
    \end{enumerate}
    
   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
  \item[]
  \item[]
                
  \item [Problem 4.15]
    \end{enumerate}
    \begin{enumerate}
        
   \item[] Suppose:
   
   \[y\;is\;N_{4}(\mu, \Sigma),\;with\]
   
   \[\mu = \begin{pmatrix} -4 \\ 
             2 \\
               5\\
               -1 \end{pmatrix}
   \Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               0 & 2 & 0 & 7 \end{pmatrix}\]
                               
    Which of the following random variables are independent?
    
    Overall Solution: Two individual variables are independent if \[\sigma_{jk}\;=\;0\]

    
    \item[(a)] \[y_{1}\;and\;y_{2}\]
    
   \[ Solution: \sigma_{12} = 0\;\;\;INDEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               \hlight{0} & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               0 & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(b)] \[y_{1}\;and\;y_{3}\]
    
    \[Solution: \sigma_{13} = -1\;\;\;DEPENDENT\]
    
      \[ \Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               \hlight{-1} & 0 & 5 & 0 \\
                               0 & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(c)] \[y_{1}\;and\;y_{4}\]
    
    \[Solution: \sigma_{14} = 0\;\;\;INDEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               \hlight{0} & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(d)] \[y_{2}\;and\;y_{3}\]
    
   \[] Solution: \sigma_{23} = 0\;\;\;INDEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & \hlight{0} & 5 & 0 \\
                               0 & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(e)] \[y_{2}\;and\;y_{4}\]
    
    \[Solution: \sigma_{24} = 2\;\;\;DEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               0 & \hlight{2} & 0 & 7 \end{pmatrix}\]
    
    
    \item[(f)] \[y_{3}\;and\;y_{4}\]
    
    \[Solution: \sigma_{34} = 0\;\;\;INDEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               0 & 2 & \hlight{0} & 7 \end{pmatrix}\]
    
    
    \item[(g)] \[(y_{1},\;y_{2})\;and\;y_{3}\]
    
    \[Solution: \sigma_{13}\;and\;\sigma_{23} = -1\;\;\;DEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               \hlight{-1} & \hlight{0} & 5 & 0 \\
                               0 & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(h)] \[(y_{1},\;y_{2})\;and\;y_{4}\]
    
   \[ Solution: \sigma_{14}\;and\;\sigma_{24} = 2\;\;\;DEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               \hlight{0} & \hlight{2} & 0 & 7 \end{pmatrix}\]
    
    
    \item[(i)] \[(y_{1},\;y_{3})\;and\;y_{4}\]
    
   \[] Solution: \sigma_{14}\;and\;\sigma_{34} = 0\;\;\;INDEPENDENT\]
    
      \[ \Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               \hlight{0} & 2 & \hlight{0} & 7 \end{pmatrix}\]
    
    
    \item[(j)] \[y_{1}\;and\;(y_{2},\;y_{4})\]
    
   \[] Solution: \sigma_{12}\;and\;\sigma_{14} = 0\;\;\;INDEPENDENT\]
    
      \[ \Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               \hlight{0} & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               \hlight{0} & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(k)] \[y_{1}\;and\;y_{2}\;and\;y_{3}\]
    
    \[Solution: \sigma_{12}\;and\;\sigma_{13};and\;\sigma_{23} = -1\;\;\;DEPENDENT\]
    
      \[ \Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               \hlight{0} & 3 & 0 & 2 \\
                               \hlight{-1} & \hlight{0} & 5 & 0 \\
                               0 & 2 & 0 & 7 \end{pmatrix}\]
    
    
    \item[(l)] \[y_{1}\;and\;y_{2}\;and\;y_{4}\]
    
    \[Solution: \sigma_{12}\;and\;\sigma_{14};and\;\sigma_{24} = 2\;\;\;DEPENDENT\]

    
  
      \[ \Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               \hlight{0} & 3 & 0 & 2 \\
                               -1 & 0 & 5 & 0 \\
                               \hlight{0} & \hlight{2} & 0 & 7 \end{pmatrix}\]
    
    
    \item[(m)] \[(y_{2},\;y_{1})\;and\;(y_{3},\;y_{4})\]
    
      \[Solution: \sigma_{23}\;and\;\sigma_{24};and\;\sigma_{13};and\;\sigma_{14} = 1\;\;\;DEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               0 & 3 & 0 & 2 \\
                               \hlight{-1} & \hlight{0} & 5 & 0 \\
                               \hlight{0} & \hlight{2} & 0 & 7 \end{pmatrix}\]
    
    
    \item[(n)] \[(y_{1},\;y_{3})\;and\;(y_{2},\;y_{4})\]
    
     \[Solution: \sigma_{12}\;and\;\sigma_{14};and\;\sigma_{32}and\;\sigma_{34} = 0\;\;\;INDEPENDENT\]
    
       \[\Sigma = \begin{pmatrix}   8 & 0 & -1 & 0 \\ 
                               \hlight{0} & 3 & \hlight{0} & 2 \\
                               -1 & 0 & 5 & 0 \\
                               \hlight{0} & 2 & \hlight{0} & 7 \end{pmatrix}\]
                               
 
 
 \end{enumerate}
 \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 4.17]
    \end{enumerate}
    \begin{enumerate}
        
   \item[] Suppose y and x are subvectors, such that y is 2 x 1 and x is 3 x 1, with the mean and Sigma partitioned accordingly:
   
     \[ \mu = \begin{pmatrix} 3 \\
                               2 \\
                               \hline
                               4 \\
                               3 \\
                               5 \\
                               \end{pmatrix} 
   \Sigma = \left ( \begin{array}{cc|ccc} 14 & -8 & 15 & 0 & 3 \\ 
   -8 & 18 &  8 & 6 & -2 \\ 
   \hline 15 & 8 & 50 & 8 & 5\\ 
   0 & 6 & 8 & 4 & 0\\ 
   3 & -2 &  5 & 0 & 1\end{array} \right ) \]
    
    Assume that:
    
    \[\frac{y}{x}\;is\;distributed\;as\;N_{5}(\mu, \Sigma)\]
        
      \setlength{\itemindent}{.5in}

     \item[(a)]  Find E(y|x)
     
     Solution:  For this solution, we use the equation:
     
     \[E(y|x) = \mu_{y} + \Sigma_{yx}\Sigma_{xx}^{-1} (x - \mu_{x})\]
     
     \[E(y|x)\;=\;\begin{pmatrix} 3 \\ -2 \end{pmatrix} + \begin{pmatrix} 15 & 0 & 3 \\ 8 & 6 & -2 \end{pmatrix}
     \begin{pmatrix} 50 & 8 & 5 \\ 8 & 4 & 0 \\ 5 & 0 & 1\end{pmatrix}^{-1} 
     \begin{pmatrix} x_{1} - 4 \\ x_{2} + 3 \\ x_{3} - 5 \end{pmatrix}\]
     
     Factoring out 
    
    \[(x - \mu_{x})\]
    
    and multiplying matrices, results in
    
    \[E(y|x) = \mu_{y} + \Sigma_{yx}\Sigma_{xx}^{-1} (x - \mu_{x})\]
     
     \[E(y|x)\;=\;\begin{pmatrix} 3 \\ -2 \end{pmatrix} + \begin{pmatrix} 15 \\ -24.5 \end{pmatrix}
     \begin{pmatrix} 0 & 0 & 3 \\ .67 & .167 & -5.33 \end{pmatrix} \begin{pmatrix} x_{1}\\ x_{2}\\ x_{3}\end{pmatrix}\]
     
     which results in
     
     \[\begin{pmatrix} -12 \\ 22 \end{pmatrix} + \begin{pmatrix} 0 & 0 & 3 \\ .67 & .167 & -5.33 \end{pmatrix} \begin{pmatrix} x_{1}\\ x_{2}\\ x_{3}\end{pmatrix}\]
    
    
     \end{enumerate}
```{r}
mu <- matrix(c(3, -2, 4, -3, 5), ncol=1)
mu_y <- matrix(c(3, -2), ncol=1)  #---mean
sigma <- matrix(c(14, -8, 15, 0, 3, -8, 18, 8, 6, -2, 15, 8, 50, 8, 5, 0, 
6, 8, 4, 0, 3, -2, 5, 0 , 1), ncol=5) #---cov matrix
sigma_yx <- matrix(c(15, 0, 3, 8, 6, -2), ncol=2)
sigma_yx <- t(sigma_yx)
sigma_xx <- matrix(c(50, 8, 5, 8, 4, 0, 5, 0, 1), ncol=3)
inverse_sigma_xx <- solve(sigma_xx)
y <- sigma_yx %*% inverse_sigma_xx
y <- round(y, digits=2)
y
``` 
\begin{enumerate}

    \item[(b)] Find cov(y|x)
    
    Solution: For this solution, we use the equation:
     
     \[cov(y|x) = \Sigma_{yy} - \Sigma_{yx}\Sigma_{xx}^{-1} \Sigma_{xy}\]
     
       \[cov(y|x) = \begin{pmatrix} 14 & -8 \\ -8 & 18 \end{pmatrix} - \begin{pmatrix} 15 & 0 & 3 \\ 8 & 6 & -2 \end{pmatrix}
     \begin{pmatrix} 50 & 8 & 5 \\ 8 & 4 & 0 \\ 5 & 0 & 1\end{pmatrix}^{-1} 
     \begin{pmatrix} 15 & 8 \\ 0 & 6 \\ 3 & 2  \end{pmatrix}\]
     
     \[cov(y|x) = \begin{pmatrix} 5 & -2 \\ -1 & 1\end{pmatrix}\]
     
         \end{enumerate}
         
```{r}
mu <- matrix(c(3, -2, 4, -3, 5), ncol=1)
mu_y <- matrix(c(3, -2), ncol=1)  #---mean
sigma <- matrix(c(14, -8, 15, 0, 3, -8, 18, 8, 6, -2, 15, 8, 50, 8, 5, 0, 
6, 8, 4, 0, 3, -2, 5, 0 , 1), ncol=5) #---cov matrix
sigma_yy <- matrix(c(14, -8, -8, 18), ncol=2)
sigma_yx <- matrix(c(15, 8, 0, 6, 3, -2), ncol=3)
sigma_xy <- matrix(c(15, 0, 3, 8, 6, -2), ncol=2)
sigma_xx <- matrix(c(50, 8, 5, 8, 4, 0, 5, 0, 1), ncol=3)
inverse_sigma_xx <- solve(sigma_xx)
covyx <- sigma_yy - sigma_yx %*% inverse_sigma_xx %*% sigma_xy
covyx
``` 
\begin{enumerate}\bfseries 

\setlength{\itemindent}{.5in}


    \item [Problem 5.3]
    \end{enumerate}
    \begin{enumerate}
    \item[] Show that 
    
    \[ t^{2}(a) = T^{2}\]
    
    Solution: By substituting
    
    \[a = S_{pl}^-1 (\bar{y}_{1} - \bar{y}_{2})\]
    
    
    into the following equation
    
    
    \[t_{2}(a) = \frac{[a'(\bar{y}_{1} - \bar{y}_{2})]^{2}}{[n_{1} + n_{2})/n_{1}n_{2}] a'S_{pl}a}\]
    
    and factoring out
    
    \[[n_{1} + n_{2})/n_{1}n_{2}]\]
    
    results in
    
    \[t^{2}(a) = \frac{n_{1}n{2}}{n_{1} + n_{2}} \frac{[(\bar{y}_{1} - \bar{y}_{2})'S_{pl}^{-1}(\bar{y}_{1} - \bar{y}_{2})]^{2}}{(\bar{y}_{1} - \bar{y}_{2})'S_{pl}^{-1}S_{pl}S_{pl}^{-1}(\bar{y}_{1} - \bar{y}_{2})} = T^{2}\]
    
    \end{enumerate}
        
    \begin{enumerate}\bfseries 

\setlength{\itemindent}{.5in}

    \item [Problem 5.12]
    \end{enumerate}
    \begin{enumerate}
    \item[] Use the probe word data in Table 3.5:
    
    \item[(a)] Test:
   
    \[H_{0}: \mu = (30,25,40,25,30)'\]
    
   Solution: y and S were first calculated from the observations, and then plugged into the following test statistic formula:
    
    \[T^{2} = n(\bar{y} - \mu_{0})'S^{-1} (\bar{y} - \mu_{0})\]
             
\end{enumerate}
```{r}
probe <- matrix(c(51,27,37,42,27,43,41,38,36,26,29,36,20,22,36,18,32,
22,21,23,31,20,50,26,41,32,33,43,36,31,27,31,25,35,17,37,34,14,35,25,
20,25,32,26,42,27,30,27,29,40,38,16,28,36,25), ncol=5)
p <- 5
v <- 10
mu0 <- matrix(c(30,25,40,25,30), ncol=1)
mean1 <- mean(probe[,1])
mean2 <- mean(probe[,2])
mean3 <- mean(probe[,3])
mean4 <- mean(probe[,4])
mean5 <- mean(probe[,5])
y = matrix(c(mean1, mean2, mean3, mean4, mean5), ncol=1)
yt = t(y)
S = var(probe)
inverseS = solve(S)
T2 = 11 * (t(mu0 - y) %*% inverseS %*% (mu0 - y))
print(paste("T2 = " , T2))
```

\begin{enumerate}
    
    \item[(b)] If Ho is rejected, test each variable separately.
   \item[]
   Solution: Based on the T2 above, Ho was rejected.  As such, each variable is tested separately below.
   
         \end{enumerate}

```{r}
t1 <- t.test(probe[,1], mu = 30)
t1 <- round(t1$statistic, digits=4)
t2 <- t.test(probe[,2], mu = 25)
t2 <- round(t2$statistic, digits=4)
t3 <- t.test(probe[,3], mu = 40)
t3 <- round(t3$statistic, digits=4)
t4 <- t.test(probe[,4], mu = 25)
t4 <- round(t4$statistic, digits=4)
t5 <- t.test(probe[,5], mu = 30)
t5 <- round(t5$statistic, digits=4)
print(paste(" t1 = " , t1 , " t2 = " , t2 , " t3 = " , t3 , " t4 = " , 
t4 , " t5 = " , t5 , sep=""))
```

\begin{enumerate}\bfseries 

\setlength{\itemindent}{.5in}

    \item [Problem 5.16]
    \end{enumerate}
    \begin{enumerate}
    \item[] Four measurements were made on two species of flea beetles.  The variables were:
    
    y1 = distance of transverse groove from posterior border of prothorax\\
    y2 = length of elytra\\
    y3 = length of second antennal joint\\
    y4 = length of third antennal joint\\
  
  \item[(a)] 
  
  \[Test\;H_{0}: \mu_{1} = \mu_{2}\;using\;T^{2}\]
  
  Solution:
   
   
                               
\end{enumerate}
   
```{r}
v <- (19 + 20 -2)
p <- 4
beetle1 <- matrix(c(189,192,217,221,171,192,213,192,170,201,195,205,180,192,200,192,200,
181,192,245,260,276,299,239,262,278,255,244,276,242,263,252,283,294,277,287,255,287,137,
132,141,142,128,147,136,128,128,146,128,147,121,138,138,150,136,146,141,163,217,192,213,
158,173,201,185,192,186,192,192,167,183,188,177,173,183,198), ncol=4)
beetle2 <- matrix(c(181,158,184,171,181,181,177,198,180,177,176,192,176,169,164,181,192,
181,175,197,305,237,300,273,297,308,301,308,286,299,317,312,285,287,265,308,276,278,271,
303,184,133,166,162,163,160,166,141,146,171,166,166,141,162,147,157,154,149,140,170,209,
188,231,213,224,223,221,197,214,192,213,209,200,214,192,204,209,235,192,205), ncol=4)
beetle1mean1 <- mean(beetle1[,1])
beetle1mean2 <- mean(beetle1[,2])
beetle1mean3 <- mean(beetle1[,3])
beetle1mean4 <- mean(beetle1[,4])
beetle2mean1 <- mean(beetle2[,1])
beetle2mean2 <- mean(beetle2[,2])
beetle2mean3 <- mean(beetle2[,3])
beetle2mean4 <- mean(beetle2[,4])
beetle1mean <- matrix(c(beetle1mean1, beetle1mean2, beetle1mean3, beetle1mean4))
beetle2mean <- matrix(c(beetle2mean1, beetle2mean2, beetle2mean3, beetle2mean4))
S1 <- cov(beetle1)
S2 <- cov(beetle2)
Spooled <- 1/(19+20-2)
Spooled2 <- 18 * S1 + 19 * S2
Spooled <- Spooled * Spooled2
InverseSpooled <- solve(Spooled)
T21 <- (19 * 20/(19 + 20)) 
T22 <- (t(beetle1mean - beetle2mean))
T23 <- (beetle1mean - beetle2mean)
T2 <- T21 * T22 %*% InverseSpooled %*% T23
T2 <- round(T2, digits=4)
print(paste("T2 = " , T2))
```
   
  Based on the T2 value above, we can reject the Ho hypothesis (Ho:mean1 = mean2).
  
   \begin{enumerate}

\setlength{\itemindent}{.5in}
                               
    \item[(b)] If the T2 test in part a rejects H0, carry out a t-test on each variable.
   
   Solution: Based on the rejection of the hypothesis above, we test each variable below: 
   
   \end{enumerate}
   
```{r}
beetleall <- matrix(c(189,192,217,221,171,192,213,192,170,201,195,205,180,192,200,192,200,
181,192,NA,245,260,276,299,239,262,278,255,244,276,242,263,252,283,294,277,287,255,287,NA,137,
132,141,142,128,147,136,128,128,146,128,147,121,138,138,150,136,146,141,NA,163,217,192,213,
158,173,201,185,192,186,192,192,167,183,188,177,173,183,198,NA,181,158,184,171,181,181,177,
198,180,177,176,192,176,169,164,181,192,181,175,197,305,237,300,273,297,308,301,308,286,299,
317,312,285,287,265,308,276,278,271,303,184,133,166,162,163,160,166,141,146,171,166,166,141,
162,147,157,154,149,140,170,209,188,231,213,224,223,221,197,214,192,213,209,200,214,192,204,
209,235,192,205), ncol=8)
t1 <- t.test(beetleall[,1], beetleall[,5])
t1 <- round(t1$statistic, digits=4)
t2 <- t.test(beetleall[,2], beetleall[,6])
t2 <- round(t2$statistic, digits=4)
t3 <- t.test(beetleall[,3], beetleall[,7])
t3 <- round(t3$statistic, digits=4)
t4 <- t.test(beetleall[,4], beetleall[,8])
t4 <- round(t4$statistic, digits=4)
print(paste(" t1 = " , t1 , " t2 = " , t2 , " t3 = " , t3 , " t4 = " , 
t4 , sep=""))
```
   
   \begin{enumerate}
   
   \setlength{\itemindent}{.5in}

                               
    \item[(c)] Calculate the discriminant function coefficient vector:
    
    \[ a = S_{pl}^{-1} (\bar{y}_{1} - \bar{y}_{2})\]
    
   Solution: 
   
   \end{enumerate}
   
```{r}
y1mean <- mean(cbind(beetleall[,1]), na.rm = TRUE)
y2mean <- mean(cbind(beetleall[,2]), na.rm = TRUE)
y3mean <- mean(cbind(beetleall[,3]), na.rm = TRUE)
y4mean <- mean(cbind(beetleall[,4]), na.rm = TRUE)
y5mean <- mean(cbind(beetleall[,5]), na.rm = TRUE)
y6mean <- mean(cbind(beetleall[,6]), na.rm = TRUE)
y7mean <- mean(cbind(beetleall[,7]), na.rm = TRUE)
y8mean <- mean(cbind(beetleall[,8]), na.rm = TRUE)
ybar1 <- c(y1mean, y2mean, y3mean, y4mean)
ybar2 <- c(y5mean, y6mean, y7mean, y8mean)
S1 <- cov(beetle1)
S2 <- cov(beetle2)
Spooled <- 1/(19+20-2)
Spooled2 <- 18 * S1 + 19 * S2
Spooled <- Spooled * Spooled2
InverseSpooled <- solve(Spooled)
a = InverseSpooled %*% (ybar1 - ybar2)
a <- round(a, digits=4)
print(paste("discriminant function coefficient vector = ", a, sep=""))
```

\begin{enumerate}

\setlength{\itemindent}{.5in}

                               
    \item[(d)] Show that if the vector a found in part (c) is substituted into t2, the result is the same as the value of T2.
    
    Solution:  Using the following equation:
    
    \[t(a) = \frac{a'\bar{y}_{1} - a'\bar{y}_{2}}{\sqrt{[(n_{1} + n_{2})/n_{1}n_{2}] a' S_{pl} a}}\]
    
   
   \end{enumerate}

```{r}
ta2_1 <- (19 + 20)/(19 * 20)
ta2_2 <- t(a) %*% Spooled %*% a
ta2_3 <- (t(a) %*% ybar1) - (t(a) %*% ybar2)
t <- ta2_3/sqrt(ta2_1 %*% ta2_3)
t2 <- t^2
t2
```
   
   \begin{enumerate}

\setlength{\itemindent}{.5in}

                               
    \item[(e)] Obtain T2 using the regression approach.
    
   Solution: 
      
   \end{enumerate}

```{r}
w2 <- -19/39
w1 <- 20/39
w <- c(-0.4871795,-0.4871795,-0.4871795,-0.4871795,-0.4871795,
-0.4871795,-0.4871795,-0.4871795,-0.4871795,-0.4871795,-0.4871795,
-0.4871795,-0.4871795,-0.4871795,-0.4871795,-0.4871795,-0.4871795,-0.4871795,
-0.4871795,0.5128205,0.5128205,0.5128205,0.5128205,0.5128205,
0.5128205,0.5128205,0.5128205,0.5128205,0.5128205,0.5128205,0.5128205,
0.5128205,0.5128205,0.5128205,0.5128205,0.5128205,0.5128205,
0.5128205,0.5128205)
w <- data.frame(w)
beetlecombined <- rbind(beetle1, beetle2)
beetlecombined <- cbind(beetlecombined, w)
colnames(beetlecombined) <- c("y1", "y2", "y3", "y4", "w")
beetlecombined.lm <- lm(w ~ y1 + y2 + y3 + y4, data=beetlecombined)
R2 <- round(summary(beetlecombined.lm)$r.squared, digits=5)
print(paste("R2 = ", R2, sep=""))
```

   
   Calculate T2 from R2 using the following formula:
   
   \[T2 = (n_{1} + n_{2} - 2)\frac{R^{2}}{1 - R^{2}}\]
   
   

```{r}
R2 <- round(summary(beetlecombined.lm)$r.squared, digits=5)
T2 <- (19 + 20 -2)* (R2/(1 - R2))
T2 <- round(T2, digits=5)
print(paste("T2 = ", T2, sep=""))
```
 
   \begin{enumerate}

\setlength{\itemindent}{.5in}

                               
    \item[(f)] Test the significance of each variable adjusted for the other three.
    
    Solution:  The following formula was used to calculate t2 for each variable adjusted for the other three.
    
    \[t^{2}(x|y) = (v - p) \frac{T_{p+1}^{2} - T_{p}^{2} }{v + T_{p}^{2}}\]

   \end{enumerate}

```{r}
beetle1 <- matrix(c(189,192,217,221,171,192,213,192,170,201,195,205,180,192,200,192,200,
181,192,245,260,276,299,239,262,278,255,244,276,242,263,252,283,294,277,287,255,287,137,
132,141,142,128,147,136,128,128,146,128,147,121,138,138,150,136,146,141,163,217,192,213,
158,173,201,185,192,186,192,192,167,183,188,177,173,183,198), ncol=4)
beetle2 <- matrix(c(181,158,184,171,181,181,177,198,180,177,176,192,176,169,164,181,192,
181,175,197,305,237,300,273,297,308,301,308,286,299,317,312,285,287,265,308,276,278,271,
303,184,133,166,162,163,160,166,141,146,171,166,166,141,162,147,157,154,149,140,170,209,
188,231,213,224,223,221,197,214,192,213,209,200,214,192,204,209,235,192,205), ncol=4)
beetle1mean1 <- mean(beetle1[,1])
beetle1mean2 <- mean(beetle1[,2])
beetle1mean3 <- mean(beetle1[,3])
beetle1mean4 <- mean(beetle1[,4])
beetle2mean1 <- mean(beetle2[,1])
beetle2mean2 <- mean(beetle2[,2])
beetle2mean3 <- mean(beetle2[,3])
beetle2mean4 <- mean(beetle2[,4])
y1beetlemean234 <- rbind(beetle1mean2, beetle1mean3, beetle1mean4)
y2beetlemean234 <- rbind(beetle2mean2, beetle2mean3, beetle2mean4)
y1beetlemean134 <- rbind(beetle1mean1, beetle1mean3, beetle1mean4)
y2beetlemean134 <- rbind(beetle2mean1, beetle2mean3, beetle2mean4)
y1beetlemean124 <- rbind(beetle1mean1, beetle1mean2, beetle1mean4)
y2beetlemean124 <- rbind(beetle2mean1, beetle2mean2, beetle2mean4)
y1beetlemean123 <- rbind(beetle1mean1, beetle1mean2, beetle1mean3)
y2beetlemean123 <- rbind(beetle2mean1, beetle2mean2, beetle2mean3)
v <- (19 + 20 - 2)
p <- 4
Syy1 <- Spooled[-c(1), -c(1)] 
InverseSyy1 <- solve(Syy1)
Syy2 <- Spooled[-c(2), -c(2)]
InverseSyy2 <- solve(Syy2)
Syy3 <- Spooled[-c(3), -c(3)]
InverseSyy3 <- solve(Syy3)
Syy4 <- Spooled[-c(4), -c(4)]
InverseSyy4 <- solve(Syy4)
T21 <- (19 * 20/(19 + 20)) 
T22 <- (t(beetle1mean - beetle2mean))
T23 <- (beetle1mean - beetle2mean)
T2 <- T21 * T22 %*% InverseSpooled %*% T23
y1t2 <- ((19 * 20)/(19 + 20)) * t(y1beetlemean234 - y2beetlemean234) %*% InverseSyy1 %*% 
(y1beetlemean234 - y2beetlemean234)
T2y1 <- (v - 3) * (T2 - y1t2)/(v + y1t2)
T2y1 <- round(T2y1, digits=4)
y2t2 <- ((19 * 20)/(19 + 20)) * t(y1beetlemean134 - y2beetlemean134) %*% InverseSyy2 %*% 
(y1beetlemean134 - y2beetlemean134)
T2y2 <- (v - 3) * (T2 - y2t2)/(v + y2t2)
T2y2 <- round(T2y2, digits=4)
y3t2 <- ((19 * 20)/(19 + 20)) * t(y2beetlemean124 - y1beetlemean124) %*% InverseSyy3 %*% 
(y2beetlemean124 - y1beetlemean124)
T2y3 <- (v - 3) * (T2 - y3t2)/(v + y3t2)
T2y3 <- round(T2y3, digits=4)
y4t2 <- ((19 * 20)/(19 + 20)) * t(y1beetlemean123 - y2beetlemean123) %*% InverseSyy4 %*% 
(y1beetlemean123 - y2beetlemean123)
T2y4 <- (v - 3) * (T2 - y4t2)/(37 + y4t2)
T2y4 <- round(T2y4, digits=4)
print(paste(" y1|y2,y3,y4 = ", T2y1, "   y2|y1,y3,y4 = ", T2y2,    
  "   y3|y1,y2,y4 = ", T2y3, "   y4|y1,y2,y3 = ", T2y4, sep=""))
```
    
   \begin{enumerate}


\setlength{\itemindent}{.5in}

             
    \item[(g)] Test the significance of y3 and y4 adjusted for y1 and y2.
    
    Solution:  The following formula was used to calculate t2 for y3 and y4 adjusted for y1 and y2.
    
        \[t^{2}(x|y) = (v - p) \frac{T_{p+1}^{2} - T_{p}^{2} }{v + T_{p}^{2}}\]
        
   \end{enumerate}

```{r}
beetle1 <- matrix(c(189,192,217,221,171,192,213,192,170,201,195,205,180,192,200,192,200,
181,192,245,260,276,299,239,262,278,255,244,276,242,263,252,283,294,277,287,255,287,137,
132,141,142,128,147,136,128,128,146,128,147,121,138,138,150,136,146,141,163,217,192,213,
158,173,201,185,192,186,192,192,167,183,188,177,173,183,198), ncol=4)
beetle2 <- matrix(c(181,158,184,171,181,181,177,198,180,177,176,192,176,169,164,181,192,
181,175,197,305,237,300,273,297,308,301,308,286,299,317,312,285,287,265,308,276,278,271,
303,184,133,166,162,163,160,166,141,146,171,166,166,141,162,147,157,154,149,140,170,209,
188,231,213,224,223,221,197,214,192,213,209,200,214,192,204,209,235,192,205), ncol=4)
beetle1mean1 <- mean(beetle1[,1])
beetle1mean2 <- mean(beetle1[,2])
beetle1mean3 <- mean(beetle1[,3])
beetle1mean4 <- mean(beetle1[,4])
beetle2mean1 <- mean(beetle2[,1])
beetle2mean2 <- mean(beetle2[,2])
beetle2mean3 <- mean(beetle2[,3])
beetle2mean4 <- mean(beetle2[,4])
y1beetlemean234 <- rbind(beetle1mean2, beetle1mean3, beetle1mean4)
y2beetlemean234 <- rbind(beetle2mean2, beetle2mean3, beetle2mean4)
y1beetlemean134 <- rbind(beetle1mean1, beetle1mean3, beetle1mean4)
y2beetlemean134 <- rbind(beetle2mean1, beetle2mean3, beetle2mean4)
y1beetlemean124 <- rbind(beetle1mean1, beetle1mean2, beetle1mean4)
y2beetlemean124 <- rbind(beetle2mean1, beetle2mean2, beetle2mean4)
y1beetlemean123 <- rbind(beetle1mean1, beetle1mean2, beetle1mean3)
y2beetlemean123 <- rbind(beetle2mean1, beetle2mean2, beetle2mean3)
y1beetlemean12 <- rbind(beetle1mean1, beetle1mean2)
y2beetlemean12 <- rbind(beetle2mean1, beetle2mean2)
y1beetlemean34 <- rbind(beetle1mean3, beetle1mean4)
y2beetlemean34 <- rbind(beetle2mean3, beetle2mean4)
v <- (19 + 20 - 2)
p <- 4
Syy12 <- Spooled[-c(3,4), -c(3,4)] 
InverseSyy12 <- solve(Syy12)
Syy34 <- Spooled[-c(1,2), -c(1,2)] 
InverseSyy34 <- solve(Syy34)
T21 <- (19 * 20/(19 + 20)) 
T22 <- (t(beetle1mean - beetle2mean))
T23 <- (beetle1mean - beetle2mean)
T2 <- T21 * T22 %*% InverseSpooled %*% T23

y12t2 <- ((19 * 20)/(19 + 20)) * t(y1beetlemean12 - y2beetlemean12) %*% InverseSyy12 %*% 
(y1beetlemean12 - y2beetlemean12)
T2y12 <- (v - 2) * (T2 - y12t2)/(v + y12t2)
T2y12 <- round(T2y12, digits=4)

print(paste("y3 and y4 | y1 and y2 =", T2y12, sep=""))
```



    