---
title: "University of Idaho Statistics 519 Spring 2015 Homework #3"
author: "Author: David Erich Seamon - erichs@uidaho.edu "
date: "Due Date: 03/13/2015"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{enumitem}
- \usepackage{tikz,amsmath}
- \newcommand\hlight[1]{\tikz[overlay, remember picture,baseline=-\the\dimexpr\fontdimen22\textfont2\relax]\node[rectangle,fill=blue!50,rounded corners,fill opacity = 0.2,draw,thick,text opacity =1] {$#1$};} 
- \pagestyle{fancy}
- \usepackage{amsmath}
- \newcommand\sufr[3][0pt]{$\rule{0pt}{\dimexpr#1+1.4ex\relax}^\frac{#2}{#3}$}
- \fancyhead[CO,CE]{Erich Seamon - University of Idaho}
- \fancyfoot[CO,CE]{Statistics 519 - Homework 3 - 03/13/2015}
- \renewcommand{\headrulewidth}{0.4pt}% Default \headrulewidth is 0.4pt
- \renewcommand{\footrulewidth}{0.4pt}% Default \footrulewidth is 0pt
- \fancyfoot[LE,RO]{\thepage}
output:
  pdf_document:
    keep_tex: yes
    
    

---

\noindent {\bf ASSIGNED PROBLEMS: 5.13, 5.21, 6.9, 6.27, 6.28, 6.31, 6.33  (plus: carry out
nonparametric hypotheses testing for parallelism, equal levels, and flatness) }

\begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}
    \item [Problem 5.13]
    
    \end{enumerate}
    \begin{enumerate}
    

    \item[] For the probe word data in Table 3.5, test:
    
    \[H_{0}: \mu_{1} = \mu_{2} = ... = \mu_{5},\;using\;T^{2}\]
    
    Solution:
  
    \end{enumerate}

```{r}
probe <- matrix(c(51,27,37,42,27,43,41,38,36,26,29,36,20,22,36,18,32,
22,21,23,31,20,50,26,41,32,33,43,36,31,27,31,25,35,17,37,34,14,35,25,
20,25,32,26,42,27,30,27,29,40,38,16,28,36,25), ncol=5)
p <- 5
v <- 10
n <- 11
C <- matrix(c(1,0,0,0,-1,1,0,0,0,-1,1,0,0,0,-1,1,0,0,0,-1), ncol=5)
Ct <- t(C)
mean1 <- mean(probe[,1])
mean2 <- mean(probe[,2])
mean3 <- mean(probe[,3])
mean4 <- mean(probe[,4])
mean5 <- mean(probe[,5])
y <- matrix(c(mean1, mean2, mean3, mean4, mean5), ncol=1)
yt <- t(y)
S <- var(probe)
inverseS <- solve(S)
t1 <- t(C %*% y)
t2 <- solve((C %*% S %*% Ct)/n)
t3 <- (C %*% y)
T2 <- t1 %*% t2 %*% t3 #---perform T2 calc using y mean vector, S, and contrast matrix
  dimx<-dim(probe)
  fval<-(dimx[1]-dimx[2])*T2/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pval<-1-pf(fval,df1,df2)
  cat("T-squared=",T2,fill=TRUE)
  df<-cbind(df1,df2)
  cat("F-value=",fval," df=",df," p-value=",pval, "  NOT SIGNIFICANT")
```

  \begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}
\item[]
    \item [Problem 5.21]
    
    \end{enumerate}
    \begin{enumerate}
    

    \item[] Each of 15 students wrote an informal and a formal essay (Kramer 1972, p. 100).  The variables recorded were the number of words and the number of verbs:
    
    \[y_{1} = number\;of\;words\;in\;the\;informal\;essay,\]
    \[y_{2} = number\;of\;verbs\;in\;the\;informal\;essay,\]
    \[y_{3} = number\;of\;words\;in\;the\;formal\;essay,\]
    \[y_{4} = number\;of\;verbs\;in\;the\;formal\;essay.\]
    
    The data are given in Table 5.9. Since each student wrote both types of essays, the observation vectors are paired, and we use the paired comparison test.
    
    \item[]
    
  
    
    \item[(a)] Test:
    
    \[H_{0}: \mu_{d} = 0\]

    Solution:
    
    In order to test the above hypothesis, we need to use the following equation:
    
    \[T = \bar{d}'\begin {pmatrix} S_{d} \\ \hline n \end{pmatrix}^{-1} \bar{d}\] 
    
    
    \end{enumerate}
    
```{r}
essay <- read.table("/nethome/erichs/multivariate_analysis/T5_9_ESSAY.DAT")
essay <- essay[,2:5]

diff1 <- matrix(c(11,-5,-80,-105,-57,-39,-35,-35,-26,3,-61,0,-40,-26,3), ncol=1)
diff2 <- matrix(c(5,-1,-8,-15,-7,-9,-1,-3,-5,4,-2,5,-6,-2,-8), ncol=1)

diff <- matrix(c(11,-5,-80,-105,-57,-39,-35,-35,-26,3,-61,0,-40,-26,
3,5,-1,-8,-15,-7,-9,-1,-3,-5,4,-2,5,-6,-2,-8), ncol=2)

C <- matrix(c(1,0,-1,0,0,-1), ncol=3)
Ct <- t(C)
p <- 2
n <- 15
v <- 14
mean1 <- mean(essay[,1])
mean2 <- mean(essay[,2])
mean3 <- mean(essay[,3])
mean4 <- mean(essay[,4])
y1 <- mean1 - mean3
y2 <- mean2 - mean4
Ta <- (n * n)/(n + n)

#--calculate dbar
dbar1 <- (1/15) * (sum(diff1))
dbar1 <- t(dbar1)
dbar2 <- (1/15) * (sum(diff2))
dbar2 <- t(dbar2)
dbar <- rbind(dbar1,dbar2)
dbart <- t(rbind(dbar1,dbar2))

#--Calculate Sd
x<- var(diff)
xx <- sqrt(x)
diffvar <- var(diff) 
diffvar <- solve(diffvar/n)

#--Calc T2
T2 <- dbart %*% diffvar %*% dbar
#print(paste("T2 = " , T2))
 dimx<-dim(essay)
  fval<-(dimx[1]-dimx[2])*T2/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pval<-1-pf(fval,df1,df2)
  cat("T-squared=",T2,fill=TRUE)
  df<-cbind(df1,df2)
  cat("F-value=",fval," df=",df," p-value=",pval, "  NOT SIGNIFICANT")
```
    
  \begin{enumerate}
    \item[]
    \item[(b)] Find the discriminant function coefficient vector.
    \item[]
    \item[]
    Solution:  
    \item[]
    
    
    we calculate the discriminant function coefficient vector with the following equation:
    
    \[a = S_{d}^{-1} \bar{d}\]
    
    \end{enumerate}
    
```{r}
essay <- read.table("/nethome/erichs/multivariate_analysis/T5_9_ESSAY.DAT")
essay <- essay[,2:5]
diff1 <- matrix(c(11,-5,-80,-105,-57,-39,-35,-35,-26,3,-61,0,-40,-26,3), ncol=1)
diff2 <- matrix(c(5,-1,-8,-15,-7,-9,-1,-3,-5,4,-2,5,-6,-2,-8), ncol=1)
diff <- matrix(c(11,-5,-80,-105,-57,-39,-35,-35,-26,3,-61,0,-40,-26,3,5,
-1,-8,-15,-7,-9,-1,-3,-5,4,-2,5,-6,-2,-8), ncol=2)
C <- matrix(c(1,0,-1,0,0,-1), ncol=3)
Ct <- t(C)
p <- 2
n <- 15
v <- 14
mean1 <- mean(essay[,1])
mean2 <- mean(essay[,2])
mean3 <- mean(essay[,3])
mean4 <- mean(essay[,4])
y1 <- mean1 - mean3
y2 <- mean2 - mean4
Ta <- (n * n)/(n + n)

#--calculate dbar
dbar1 <- (1/15) * (sum(diff1))
dbar1 <- t(dbar1)
dbar2 <- (1/15) * (sum(diff2))
dbar2 <- t(dbar2)
dbar <- rbind(dbar1,dbar2)
dbart <- t(rbind(dbar1,dbar2))

#--Calculate Sd
x<- var(diff)
xx <- sqrt(x)
diffvar <- var(diff)
diffvar2 <- diffvar
diffvar <- solve(diffvar/n)

#---discriminant function coef vector
a <- solve(diffvar2) %*% dbar 
a
```
    
  \begin{enumerate}
    \item[]
    
    \item[(c)] Do a univariate t-test on each dj.
    \item[]
    
    Solution:
    \item[]
    
    \end{enumerate}

```{r}
t1  <- t.test(diff[,1])
t1a <- round(t1$statistic, digits=4)
t1p <- round(t1$p.value, digits=4)
t2  <- t.test(diff[,2])
t2a <- round(t2$statistic, digits=4)
t2p <- round(t2$p.value, digits=4)
cat(paste(" t1 = " , t1a , "  pvalue = ", t1p,  "   SIGNIFICANT", "\n", " t2 = " , 
t2a, "  pvalue = ", t2p, "   SIGNIFICANT", sep=""))
```

  \begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}
\item[]
\item[]
    \item [Problem 6.9]
    
    \end{enumerate}
    \begin{enumerate}
   
   \item[] Show that for two groups:
   
    \[H = \sum\limits_{i=1}^{2} n_{i}(\bar{y}_{i.} - \bar{y}_{..})(\bar{y}_{i.} - \bar{y}_{..})'\]
               
               can be expressed as:
               
    \[H = [n_{1}n_{2}/(n_{1} + n_{2})](\bar{y}_{1.} - \bar{y}_{..})(\bar{y}_{1.} - \bar{y}_{2.})'\]
     
               Solution:
               
               Assuming that:
               \[\bar{y}_{..} = \frac{n_{1}\bar{y}_{1.} + n_{2}\bar{y}_{2.}}{n_{1} + n_{2}}\]
               
               then the above can be substituted for ybar, resulting in:
               
               \[\bar{y}_{1.} - \frac{n_{1}\bar{y}_{1.} + n_{2}\bar{y}_{2.}}{n_{1} + n_{2}} = \frac{n_{2}(\bar{y}_{1.} - \bar{y}_{2.}}{n_{1} + n_{2}}\]
               
               For two groups (n=2), if we plug in the above, results in:
               
               \[\frac{n_{1}n_{2}^{2}}{(n_{1} + n_{2})^{2}}(\bar{y}_{1.} - \bar{y}_{2.})(\bar{y}_{1.} - \bar{y}_{2.})'\;and\;\frac{n_{1}^{2}n_{2}}{(n_{1} + n_{2})^{2}}(\bar{y}_{1.} - \bar{y}_{2.})(\bar{y}_{1.} - \bar{y}_{2.})'\]

    \end{enumerate}

   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 6.27]
    \end{enumerate}
    \begin{enumerate}

        \item[(a)] Compare the three methods using all four MANOVA tests.
        
        Solution:
        
    \end{enumerate}

```{r}
fish <- read.table("/nethome/erichs/multivariate_analysis/T6_17_FISH.DAT")
#--factorize column 1
p <- 4
n <- 36
v <- 35
fish1 <- fish[1:12,]
fish2 <- fish[13:24,]
fish3 <- fish[25:36,]
fishfactor <- factor(fish[,1])
fish <- cbind(fish[2:5])
fish <- as.matrix(fish)
#method <- factor(gl(3,18,36,labels=c("M1","M2","M3")))
#--Run manovas
sumfish <- summary(manova(fish ~ fishfactor))
sumfish_Wilks <- summary(manova(fish ~ fishfactor),test="W")
sumfish_Hotelling <- summary(manova(fish ~ fishfactor),test="H")
sumfish_Roy <- summary(manova(fish ~ fishfactor),test="R")
sumfish_Pillai <- summary(manova(fish ~ fishfactor),test="P")
sumfishH <- matrix(c(1.0505556,2.1733333,-1.3755556,-0.7602778,2.173333,4.880000,
-2.373333,-1.256667,-1.375556,-2.373333,2.382222,1.384444,-0.7602778,-1.2566667,
1.3844444,0.8105556), ncol=4)
sumfishE <- matrix(c(13.408333,7.723333,8.675000,5.864167,7.723333,8.480000,
7.526667,6.213333,8.675000,7.526667,11.607500,7.037500,5.864167,6.213333,
7.037500,10.565833), ncol=4)
inversesumfishE <- solve(sumfishE)
eh <- inversesumfishE %*% sumfishH
customRoy <- sumfish$Eigenvalues[,1]/(1 + sumfish$Eigenvalues[,1])
#--Pvalue calcs
print(sumfish$SS)
pvalue <- 1 - pf((v-p+1)/(v*p)+customRoy, df1=p, df2=v-p+1)
cat(paste(" Wilks = " , sumfish_Wilks$stats[3] , " P-value = ", sumfish_Wilks$stats[11], 
" - Significant", "\n", " Hotelling = " , sumfish_Hotelling$stats[3] , 
"  P-value = ", sumfish_Hotelling$stats[11], " - Significant", "\n", 
" Roy = " , sumfish_Roy$stats[3] , "  P-value = ", sumfish_Roy$stats[11], 
" - Significant", "\n", 
" Pillai = " , sumfish_Pillai$stats[3] , "  P-value = ", sumfish_Pillai$stats[11], 
" - Significant", "\n", " Corrected Roy = " , customRoy , "  P-value = ", 
sumfish_Roy$stats[11], " - Significant", "\n",sep=""))
```
  
   \setlength{\itemindent}{.5in}
                
  (b) Compute the following measures of multivariate association: 
  
  \[\eta_{\Lambda}^{2},\;\eta_{\theta}^{2},\;A_{\Lambda},\;A_{LH},\;A_{P}\]
  
  Solution:
  
  \[\eta_{\Lambda}^{2} = 1 - \Lambda = 1 - .224 = .776\]
  \[\eta_{\theta}^{2} = \theta = .746\]
  \[A_{\Lambda} = 1 - \Lambda^{1/s} = 1 - \Lambda^{1/2} = .526\]
  \[A_{LH} = \frac{U^{(s)}/s}{1 + U^{(s)}/s} = \frac{3.08/2}{1 + 3.08/2} = .606\]


(c) Based on the eigenvalues, is the essential dimensionality of the space containing the mean vectors equal to 1 or 2?

Solution:

The positive eigenvalues of the E-1H matrix are 2.95 and .127 - listed below.  As such, the dimensionality is 1.

```{r}
sumfish$Eigenvalues
```
          
(d) Using contrasts, test the followng two comparison methods: 1 and 2 vs. 3, and 1 vs. 2.


```{r}
fishy11mean <- mean(fish1[,2])
fishy12mean <- mean(fish1[,3])
fishy13mean <- mean(fish1[,4])
fishy14mean <- mean(fish1[,5])
fishy21mean <- mean(fish2[,2])
fishy22mean <- mean(fish2[,3])
fishy23mean <- mean(fish2[,4])
fishy24mean <- mean(fish2[,5])
fishy31mean <- mean(fish3[,2])
fishy32mean <- mean(fish3[,3])
fishy33mean <- mean(fish3[,4])
fishy34mean <- mean(fish3[,5])
fishy1mean <- c(fishy11mean, fishy12mean, fishy13mean, fishy14mean)
fishy2mean <- c(fishy21mean, fishy22mean, fishy23mean, fishy24mean)
fishy3mean <- c(fishy31mean, fishy32mean, fishy33mean, fishy34mean)
n <- 36
v <- 35
N <- (33 - 4 - 1)/2
m <- ((2 - 4)  - 1)/2
Ve <- 3 * (12 - 1)
Vh <- 3 - 1
p <- 4
s <- min(Vh, p)

cy <- (1 * fishy1mean) + (1 * fishy2mean) - 2 * (fishy3mean)
cy <- t(cy)
cy <- t(cy)
cy2 <- (1 * fishy1mean) - (fishy2mean)
cy2 <- t(cy2)
cy2 <- t(cy2)
df1=p
df2=v-p+1
H1 <- (12/6) * cy %*% t(cy)
H1 <- (12/6) * ((1 * fishy1mean) + (1 * fishy2mean) - 2 * (fishy3mean)) %*% 
t(((1 * fishy1mean) + (1 * fishy2mean) - 2 * (fishy3mean)))
WilksH1 <- det(sumfishE)/det(sumfishE + H1)
H1F <- ((N - p)/p*(N-1))*WilksH1
H1P <- pf(n, df1, df2, H1F, lower.tail=F) - (1-pf(36, 4, 32))
eh1 <- solve(sumfishE) %*% H1
eigeneh <- eigen(eh1)
RoyH1 <- eigeneh$values[1]/(1 + eigeneh$values[1])
RoyH1pvalue <- 1 - pf((v-p+1)/(v*p)+RoyH1, df1=p, df2=v-p+1)
PillaiH1 <- eigeneh$values[1]/(1 + eigeneh$values[1])
PillaiH1pvalue <- 1 - pf((v-p+1)/(v*p)+PillaiH1, df1=p, df2=v-p+1)
HotellingH1 <- sum(diag(eh1))
HotellingH1pvalue <- 1 - pf((v-p+1)/(v*p)+HotellingH1, df1=p, df2=v-p+1)
N <- (33 - 4 - 1)/2
m <- ((2 - 4)  - 1)/2 # Vh is 2 and p = 4 - so m = 0 is used in table lookup
Ve <- 3 * (12 - 1)
Vh <- 3 - 1
p <- 4
s <- min(Vh, p)


#--paste values
cat(paste("Methods 1 and 2 vs 3", "\n", " WilksH1 = " , 
WilksH1 ,  "   pvalue = ", H1P, "  SIGNIFICANT", "\n", 
" RoyH1 = " , RoyH1 ,  "\n", 
" PillaiH1 = " , PillaiH1 ,  "\n", 
" HotellingH1 = " , HotellingH1, sep=""))


H2 <- (12/2) * cy2 %*% t(cy2)
WilksH2 <- det(sumfishE)/det(sumfishE + H2)
H2F <- ((N - p)/p*(N-1))*WilksH2
H2P <- pf(n, df1, df2, H2F, lower.tail=F) - (1-pf(36, 4, 32))
eh2<- solve(sumfishE) %*% H2
eigeneh2 <- eigen(eh2)
RoyH2 <- eigeneh2$values[1]/(1 + eigeneh2$values[1])
PillaiH2 <- eigeneh2$values[1]/(1 + eigeneh2$values[1])
HotellingH2 <- sum(diag(eh2))
#--paste values



cat(paste("Methods 1 vs. 2", "\n", " WilksH2 = " , WilksH2 , 
"   pvalue = ", H2P, "  SIGNIFICANT", "\n", 
" RoyH2 = " , RoyH2 , "\n", " PillaiH2 = " , PillaiH2 , "\n", 
" HotellingH2 = " , HotellingH2 , sep=""))
```
        

  \setlength{\itemindent}{.5in}

        
(e) If any of the four tests in (a) is significant, run an ANOVA F-test on each yi and examine the discriminant function.
   
   Solution:
  
  

```{r}
newfish <- data.frame(fish)
aovfish <- aov(fish ~ fishfactor, data=newfish)
summary(aovfish)

#---discriminant function calculation
eh_eigenvector <- eigen(eh)
eh_eigenvector <- eh_eigenvector$vectors
api <- t(eh_eigenvector)
api1 <- api[1,]
#---print a'
api1 
```
 

 \setlength{\itemindent}{.5in}
        
   (f) Test the significance of y3 and y4 adjusted for y1 and y2.
   
   Solution:
   

```{r}
y12W <- summary(manova(fish[,1:2] ~ fishfactor), test="W")
y1234W <- summary(manova(fish ~ fishfactor),test="W")
sigtest <- y1234W$stats[3]/y12W$stats[3]
dimx <- dim(fish)
fval<-(dimx[1]-dimx[2])*sigtest/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pval<-pf(fval,df1,df2)
  cat(paste("Testing Significance of y3 and y4 adjusted for y1 and y2", 
  "\n", "T-squared=",sigtest))
  df<-cbind(df1,df2)
  cat("F-value=",fval," df=",df," p-value=",pval,  "  SIGNIFICANT")
```
 
 
 \setlength{\itemindent}{.5in}
        
   (g) Test the significance of each variable adjusted for the other three.
   
   Solution:
   
```{r}
y234W <- summary(manova(fish[,c(2,3,4)] ~ fishfactor), test="W")
y134W <- summary(manova(fish[,c(1,3,4)] ~ fishfactor), test="W")
y124W <- summary(manova(fish[,c(1,2,4)] ~ fishfactor), test="W")
y123W <- summary(manova(fish[,c(1,2,3)] ~ fishfactor), test="W")
y1234W <- summary(manova(fish ~ fishfactor),test="W")
y1adjust <- y1234W$stats[3]/y234W$stats[3]

  dimx <- dim(fish)
  fvaly1<-(dimx[1]-dimx[2])*y1adjust/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvaly1<- pf(fvaly1,df1,df2)
  df<-cbind(df1,df2)

y1adjustF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * y1adjust
y1adjustP <- pf(y1adjustF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
y2adjust <- y1234W$stats[3]/y134W$stats[3]

  dimx <- dim(fish)
  fvaly2<-(dimx[1]-dimx[2])*y2adjust/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvaly2<- pf(fvaly2,df1,df2)
  df<-cbind(df1,df2)

y2adjustF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * y2adjust
y2adjustP <- pf(y2adjustF, 2, 62, lower.tail=F) #- (1-pf(y2adjust, 2, 36))
y3adjust <- y1234W$stats[3]/y124W$stats[3]

  dimx <- dim(fish)
  fvaly3<-(dimx[1]-dimx[2])*y3adjust/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvaly3<- pf(fvaly3,df1,df2)
  df<-cbind(df1,df2)

y3adjustF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * y3adjust
y3adjustP <- pf(y3adjustF, 2, 30, lower.tail=F) #- (1-pf(36, 2, 30))
y4adjust <- y1234W$stats[3]/y123W$stats[3]

  dimx <- dim(fish)
  fvaly4<-(dimx[1]-dimx[2])*y4adjust/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvaly4<-pf(fvaly4,df1,df2)
  df<-cbind(df1,df2)

y4adjustF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * y4adjust
y4adjustP <- pf(y4adjustF, 2, 30, lower.tail=F) #- (1-pf(36, 2, 30))
cat(paste(" y1 | y2, y3, y4  = " , y1adjust , " pvalue = ", pvaly1, 
"  NOT SIGNIFICANT ", "\n", " y2 | y1, y3, y4 =  " , y2adjust , 
"  pvalue = ", pvaly2,   "  SIGNIFICANT ", "\n", 
" y3 | y1, y2, y4 =  " , y3adjust , " pvalue = ", pvaly3, 
" SIGNIFICANT ", "\n", " y4 | y1, y2, y3 =  " , y4adjust , 
" pvalue =  ", pvaly4, "  NOT SIGNIFICANT ", "\n", sep=""))
```   
 
   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 6.28]
    \end{enumerate}
    \begin{enumerate}
        
   \item[] Table 6.18 gives data from a two-way fixed effects manova on snap beans shwing the results of four variables: y1 = yield earliness, y2 = specific leaf area earliness, y3 = total yield, and y4 = average specific leaf area.  The factors are sowing date (S) and variety (V).
   \\
   
   (a) Test for main effects and interaction using all four MANOVA statistics.
   
   Solution:
     
       \end{enumerate}

```{r}
snapbean <- read.table("/nethome/erichs/multivariate_analysis/T6_18_SNAPBEAN.DAT")
colnames(snapbean) <- c("S", "V", "obs", "y1", "y2", "y3", "y4")
snapbean1 <- as.matrix(cbind(snapbean[,4:7]))
snapbeanSfactor <- factor(snapbean[,1])
snapbeanVfactor <- factor(snapbean[,2])
effectW <- summary(manova(snapbean1 ~ snapbeanSfactor*snapbeanVfactor), test="W")
effectH <- summary(manova(snapbean1 ~ snapbeanSfactor*snapbeanVfactor), test="H")
effectR <- summary(manova(snapbean1 ~ snapbeanSfactor*snapbeanVfactor), test="R")
effectP <- summary(manova(snapbean1 ~ snapbeanSfactor*snapbeanVfactor), test="P")
effectR_corrected <- effectW$Eigenvalues[,1]/(1 + effectW$Eigenvalues[,1])


effectW
effectH
effectR
effectR_corrected
effectP
```

NOTE: as listed above via the respective pvalues - all outcomes are SIGNIFICANT.

 
 \begin{enumerate}
     
\setlength{\itemindent}{.5in}
    \item[]

    \item[(b)] Compare variety 2 with varieties 1 and 3 by means of a test on a contrast.
    \item[]
    
    Solution
    \end{enumerate}

```{r}
snapbean <- read.table("/nethome/erichs/multivariate_analysis/T6_18_SNAPBEAN.DAT")
snapbean1 <- as.matrix(snapbean[4:7])
snap1 <- snapbean[c(1:5,16:20,31:35,46:50),]
snap2 <- snapbean[c(6:10,21:25,36:40,51:55),]
snap3 <- snapbean[c(11:15,26:30,41:45,56:60),]

snapall <- rbind(snap1,snap2,snap3)

snapbeanVfactor <- factor(snapall[,2])
sumsnap <- summary(manova(snapbean1 ~ snapbeanVfactor), test="W")
sumsnapE <- sumsnap$SS$Residuals

snapy11mean <- mean(snap1[,4])
snapy12mean <- mean(snap1[,5])
snapy13mean <- mean(snap1[,6])
snapy14mean <- mean(snap1[,7])

snapy21mean <- mean(snap2[,4])
snapy22mean <- mean(snap2[,5])
snapy23mean <- mean(snap2[,6])
snapy24mean <- mean(snap2[,7])

snapy31mean <- mean(snap3[,4])
snapy32mean <- mean(snap3[,5])
snapy33mean <- mean(snap3[,6])
snapy34mean <- mean(snap3[,7])

snapy1mean <- c(snapy11mean, snapy12mean, snapy13mean, snapy14mean)
snapy2mean <- c(snapy21mean, snapy22mean, snapy23mean, snapy24mean)
snapy3mean <- c(snapy31mean, snapy32mean, snapy33mean, snapy34mean)

cy <- (1 * snapy1mean) + (-2 * snapy2mean) + (1 * snapy3mean)
cy <- t(cy)
cy <- t(cy)

H1 <- (20/6) * (cy %*% t(cy))
H1 <- (20/6) * ((1 * snapy1mean) + (-2 * snapy2mean) + (1 * snapy3mean)) %*% 
t(((1 * snapy1mean) + (-2 * snapy2mean) + (1 * snapy3mean)))
snapeh <- solve(sumsnapE) %*% H1 #--E-1H

WilksH1 <- (det(sumsnapE))/(det(sumsnapE + H1)) #--creates wilks lambda
  dimx<-dim(snapbean1)
  fval1<-(dimx[1]-dimx[2])*WilksH1/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pval1<-pf(fval1,df1,df2)
  df<-cbind(df1,df2)

eh1 <- solve(sumsnapE) %*% H1
eigeneh <- eigen(eh1)
RoyH1 <- eigeneh$values[1]/(1 + eigeneh$values[1])
RoyH1pvalue <- 1 - pf((v-p+1)/(v*p)+RoyH1, df1=p, df2=v-p+1)

customRoy <- sumsnap$Eigenvalues[,1]/(1 + sumfish$Eigenvalues[,1])

PillaiH1 <- eigeneh$values[1]/(1 + eigeneh$values[1])
PillaiH1pvalue <- 1 - pf((v-p+1)/(v*p)+PillaiH1, df1=p, df2=v-p+1)
HotellingH1 <- sum(diag(eh1))
HotellingH1pvalue <- 1 - pf((v-p+1)/(v*p)+HotellingH1, df1=p, df2=v-p+1)
N <- (33 - 4 - 1)/2
m <- ((2 - 4)  - 1)/2 # Vh is 2 and p = 4 - so m = 0 is used in table lookup
Ve <- 3 * (12 - 1)
Vh <- 3 - 1
p <- 4
s <- min(Vh, p)
#--paste values
cat(paste(" WilksH1 = " , WilksH1 ,  " pvalue = ", pval1, " SIGNIFICANT ", "\n", 
" RoyH1 = " , RoyH1 , "\n", " PillaiH1 = " , PillaiH1 , "\n", " HotellingH1 = " , 
HotellingH1 , sep=""))
```
     
  \setlength{\itemindent}{.5in}
        
   (c) Test linear, quadratic, and cubic contrasts for sowing date.
   
   Solution:
   
```{r}
snapbean <- read.table("/nethome/erichs/multivariate_analysis/T6_18_SNAPBEAN.DAT")
snapbean1 <- as.matrix(snapbean[,4:7])

N <- (33 - 4 - 1)/2
m <- ((2 - 4)  - 1)/2 # Vh is 2 and p = 4 - so m = 0 is used in table lookup
Ve <- 3 * (12 - 1)
Vh <- 3 - 1
p <- 4
s <- min(Vh, p)

snap1 <- snapbean[c(1:15),]
snap2 <- snapbean[c(16:30),]
snap3 <- snapbean[c(31:45),]
snap4 <- snapbean[c(46:60),]
snapbeanSfactor <- factor(snapbean[,1])
snapbeanVfactor <- factor(snapbean[,2])
sumsnap <- summary(manova(snapbean1 ~ snapbeanSfactor), test="W")
sumsnapE <- sumsnap$SS$Residuals
snapy11mean <- mean(snap1[,4])
snapy12mean <- mean(snap1[,5])
snapy13mean <- mean(snap1[,6])
snapy14mean <- mean(snap1[,7])

snapy21mean <- mean(snap2[,4])
snapy22mean <- mean(snap2[,5])
snapy23mean <- mean(snap2[,6])
snapy24mean <- mean(snap2[,7])

snapy31mean <- mean(snap3[,4])
snapy32mean <- mean(snap3[,5])
snapy33mean <- mean(snap3[,6])
snapy34mean <- mean(snap3[,7])

snapy41mean <- mean(snap4[,4])
snapy42mean <- mean(snap4[,5])
snapy43mean <- mean(snap4[,6])
snapy44mean <- mean(snap4[,7])

snapy1mean <- c(snapy11mean, snapy12mean, snapy13mean, snapy14mean)
snapy2mean <- c(snapy21mean, snapy22mean, snapy23mean, snapy24mean)
snapy3mean <- c(snapy31mean, snapy32mean, snapy33mean, snapy34mean)
snapy4mean <- c(snapy41mean, snapy42mean, snapy43mean, snapy44mean)

cy1 <- (-3 * snapy1mean) + (-1 * snapy2mean) + (1 * snapy3mean) + (3 * snapy4mean)
cy1 <- t(cy1)
cy1 <- t(cy1)
cy2 <- (1 * snapy1mean) + (-1 * snapy2mean) + (-1 * snapy3mean) + (1 * snapy4mean)
cy2 <- t(cy2)
cy2 <- t(cy2)
cy3 <- (-1 * snapy1mean) + (3 * snapy2mean) + (-3 * snapy3mean) + (1 * snapy4mean)
cy3 <- t(cy3)
cy3 <- t(cy3)
H1 <- (15) * (cy1 %*% t(cy1))
snapeh1 <- solve(sumsnapE) %*% H1
WilksH1 <- det(sumsnapE)/det(sumsnapE + H1)
WilksH1F <- ((12 + 140 -4 -1)/(12 + 140 -2) * p) * WilksH1
WilksH1P <- 1 - pf(WilksH1F, 12, 140, lower.tail=F) #- (1-pf(36, 2, 30))

  dimx<-dim(snapbean1)
  fvalb1<-(dimx[1]-dimx[2])*WilksH1/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvalb1<-pf(fvalb1,df1,df2)
  #cat("T-squared=",WilksH1,fill=TRUE)
  df<-cbind(df1,df2)
  #cat("F-value=",fvalb1," df=",df," p-value=",pvalb1, "  NOT SIGNIFICANT")

H2 <- (15) * (cy2 %*% t(cy2))
snapeh2 <- solve(sumsnapE) %*% H2
WilksH2 <- det(sumsnapE)/det(sumsnapE + H2)
WilksH2F <- ((12 + 140 -4 -1)/(12 + 140 -2) * p) * WilksH2
WilksH2P <- 1 - pf(WilksH2F, 12, 140, lower.tail=F) #- (1-pf(36, 2, 30))

  dimx<-dim(snapbean1)
  fvalb2<-(dimx[1]-dimx[2])*WilksH2/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvalb2<-pf(fvalb2,df1,df2)
  #cat("T-squared=",WilksH2,fill=TRUE)
  df<-cbind(df1,df2)

H3 <- (15) * (cy3 %*% t(cy3))
snapeh3 <- solve(sumsnapE) %*% H3
WilksH3 <- det(sumsnapE)/det(sumsnapE + H3)
WilksH3F <- ((12 + 140 -4 -1)/(12 + 140 -2) * p) * WilksH3
WilksH3P <- 1 - pf(WilksH3F, 12, 140, lower.tail=F) #- (1-pf(36, 2, 30))

  dimx<-dim(snapbean1)
  fvalb3<-(dimx[1]-dimx[2])*WilksH3/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvalb3<-pf(fvalb3,df1,df2)
  #cat("T-squared=",WilksH3,fill=TRUE)
  df<-cbind(df1,df2)

snapeh1 <- solve(sumsnapE) %*% H1
eigeneh1 <- eigen(snapeh1)
RoyH1 <- eigeneh1$values[1]/(1 + eigeneh1$values[1])
RoyH1pvalue <- 1 - pf((v-p+1)/(v*p)+RoyH1, df1=p, df2=v-p+1)

snapeh2 <- solve(sumsnapE) %*% H2
eigeneh2 <- eigen(snapeh2)
RoyH2 <- eigeneh2$values[1]/(1 + eigeneh2$values[1])

snapeh3 <- solve(sumsnapE) %*% H3
eigeneh3 <- eigen(snapeh3)
RoyH3 <- eigeneh3$values[1]/(1 + eigeneh3$values[1])
#customRoy1 <- sumsnap$Eigenvalues[,1]/(1 + sumsnap$Eigenvalues[,1])
PillaiH1 <- eigeneh1$values[1]/(1 + eigeneh1$values[1])
PillaiH2 <- eigeneh2$values[1]/(1 + eigeneh2$values[1])
PillaiH3 <- eigeneh3$values[1]/(1 + eigeneh3$values[1])
HotellingH1 <- diag(snapeh1)
HotellingH1 <- sum(HotellingH1)
HotellingH2 <- diag(snapeh2)
HotellingH2 <- sum(HotellingH2)
HotellingH3 <- diag(snapeh3)
HotellingH3 <- sum(HotellingH3)

#--paste values
cat(paste(" LINEAR: WilksH1 = " , WilksH1 ,  " pvalue = ", pvalb1, 
" SIGNIFICANT ", "\n", " RoyH1 = " , RoyH1 , "\n", " PillaiH1 = " , 
PillaiH1 , "\n", " HotellingH1 = " , HotellingH1 , "\n", "\n", 
" QUADRATIC: WilksH2 = " , WilksH2 ,  " pvalue = ", pvalb2, " SIGNIFICANT ", 
"\n", " RoyH2 = " , RoyH2 , "\n", " PillaiH2 = " , PillaiH2 , "\n", 
" HotellingH2 = " , HotellingH2 , "\n", "\n", " CUBIC: WilksH3 = " , WilksH3 ,  
" pvalue = ", pvalb3, " SIGNIFICANT ", "\n", " RoyH3 = " , RoyH3 , "\n", 
" PillaiH3 = " , PillaiH3 , "\n", " HotellingH3 = " , HotellingH3 , "\n", 
"\n", sep=""))
```
    
 
  \setlength{\itemindent}{.5in}
        
   (d) If any of the tests in part (a) rejects Ho, carry out ANOVA F-tests on the four variables.
   
   Solution:
   

```{r}
newsnap <- data.frame(snapbean1)
aovsnapSV <- aov(snapbean1 ~ snapbeanSfactor*snapbeanVfactor, data=newsnap)
summary(aovsnapSV)
```

NOTE: as listed above, all outcomes are significant, EXCEPT the last F value for Response on V7, snapbeanSfactor:snapbeanVfactor, - which has an F of 2.08 and a pvalue of .0724.
 
  \setlength{\itemindent}{.5in}
        
   (e) Test the significance of y3 and y4 adjusted for y1 and y2 in the main effects and interaction.
   
   Solution: 
   
   
```{r}
y12W <- summary(manova(snapbean1[,1:2] ~ snapbeanSfactor*snapbeanVfactor), test="W")
Wilks12S<- y12W$stats[5]
Wilks12V<- y12W$stats[6]
Wilks12SV<- y12W$stats[7]
y1234W <- summary(manova(snapbean1 ~ snapbeanSfactor*snapbeanVfactor),test="W")
Wilks1234S<- y1234W$stats[5]
Wilks1234V<- y1234W$stats[6]
Wilks1234SV<- y1234W$stats[7]

sigtestS <- y1234W$stats[5]/y12W$stats[5]
sigtestSF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtestS
sigtestSP <- pf(sigtestSF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))

sigtestV <- y1234W$stats[6]/y12W$stats[6]
sigtestVF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtestV
sigtestVP <- pf(sigtestVF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))

sigtestSV <- y1234W$stats[7]/y12W$stats[7]
sigtestSVF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtestSV
sigtestSVP <- pf(sigtestSVF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))

cat(paste(" S - y3, y4 | y1, y2  = " , sigtestS , " pvalue = ", sigtestSP, 
" NOT SIGNIFICANT ", "\n", " V - y3, y4 | y1, y2  = " , sigtestV , 
" pvalue = ", sigtestVP, " SIGNIFICANT ", "\n", " SV - y3, y4 | y1, y2  = " , 
sigtestSV , " pvalue = ", sigtestSVP, " NOT SIGNIFICANT ", "\n", sep=""))
```
  
  
 \setlength{\itemindent}{.5in}
        
   (e) Test significance of each variable adjusted for the other three.
   
   Solution:  
   
```{r}
y234W <- summary(manova(snapbean1[,c(2,3,4)] ~ snapbeanSfactor*snapbeanVfactor), test="W")
Wilks234S<- y234W$stats[5]
Wilks234V<- y234W$stats[6]
Wilks234SV<- y234W$stats[7]
y134W <- summary(manova(snapbean1[,c(1,3,4)] ~ snapbeanSfactor*snapbeanVfactor),test="W")
Wilks134S<- y134W$stats[5]
Wilks134V<- y134W$stats[6]
Wilks134SV<- y134W$stats[7]
y124W <- summary(manova(snapbean1[,c(1,2,4)] ~ snapbeanSfactor*snapbeanVfactor),test="W")
Wilks124S<- y124W$stats[5]
Wilks124V<- y124W$stats[6]
Wilks124SV<- y124W$stats[7]
y123W <- summary(manova(snapbean1[,c(1,2,3)] ~ snapbeanSfactor*snapbeanVfactor),test="W")
Wilks123S<- y123W$stats[5]
Wilks123V<- y123W$stats[6]
Wilks123SV<- y123W$stats[7]

sigtesty1S <- y1234W$stats[5]/y234W$stats[5]
sigtesty1SF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty1S
sigtesty1SP <- pf(sigtesty1SF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty1S <- round(sigtesty1S, digits=4)
sigtesty1SP <- round(sigtesty1SP, digits=4)

sigtesty1V <- y1234W$stats[6]/y234W$stats[6]
sigtesty1VF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty1V
sigtesty1VP <- pf(sigtesty1VF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty1V <- round(sigtesty1V, digits=4)
sigtesty1VP <- round(sigtesty1VP, digits=4)

sigtesty1SV <- y1234W$stats[7]/y234W$stats[7]
sigtesty1SVF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty1SV
sigtesty1SVP <- pf(sigtesty1SVF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty1SV <- round(sigtesty1SV, digits=4)
sigtesty1SVP <- round(sigtesty1SVP, digits=4)


sigtesty2S <- y1234W$stats[5]/y134W$stats[5]
sigtesty2SF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty2S
sigtesty2SP <- pf(sigtesty2SF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty2S <- round(sigtesty2S, digits=4)
sigtesty2SP <- round(sigtesty2SP, digits=4)

sigtesty2V <- y1234W$stats[6]/y134W$stats[6]
sigtesty2VF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty2V
sigtesty2VP <- pf(sigtesty2VF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty2V <- round(sigtesty2V, digits=4)
sigtesty2VP <- round(sigtesty2VP, digits=4)

sigtesty2SV <- y1234W$stats[7]/y134W$stats[7]
sigtesty2SVF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty2SV
sigtesty2SVP <- pf(sigtesty2SVF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty2SV <- round(sigtesty2SV, digits=4)
sigtesty2SVP <- round(sigtesty2SVP, digits=4)

sigtesty3S <- y1234W$stats[5]/y124W$stats[5]
sigtesty3SF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty3S
sigtesty3SP <- pf(sigtesty3SF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty3S <- round(sigtesty3S, digits=4)
sigtesty3SP <- round(sigtesty3SP, digits=4)

sigtesty3V <- y1234W$stats[6]/y124W$stats[6]
sigtesty3VF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty3V
sigtesty3VP <- pf(sigtesty3VF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty3V <- round(sigtesty3V, digits=4)
sigtesty3VP <- round(sigtesty3VP, digits=4)

sigtesty3SV <- y1234W$stats[7]/y124W$stats[7]
sigtesty3SVF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty3SV
sigtesty3SVP <- pf(sigtesty3SVF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty3SV <- round(sigtesty3SV, digits=4)
sigtesty3SVP <- round(sigtesty3SVP, digits=4)

sigtesty4S <- y1234W$stats[5]/y123W$stats[5]
sigtesty4SF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty4S
sigtesty4SP <- pf(sigtesty4SF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty4S <- round(sigtesty4S, digits=4)
sigtesty4SP <- round(sigtesty4SP, digits=4)

sigtesty4V <- y1234W$stats[6]/y123W$stats[6]
sigtesty4VF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty4V
sigtesty4VP <- pf(sigtesty4VF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty4V <- round(sigtesty4V, digits=4)
sigtesty4VP <- round(sigtesty4VP, digits=4)

sigtesty4SV <- y1234W$stats[7]/y123W$stats[7]
sigtesty4SVF <- ((2 + 62 -4 -1)/(2 + 62 -2) * p) * sigtesty4SV
sigtesty4SVP <- pf(sigtesty4SVF, 2, 62, lower.tail=F) #- (1-pf(36, 2, 30))
sigtesty4SV <- round(sigtesty4SV, digits=4)
sigtesty4SVP <- round(sigtesty4SVP, digits=4)


cat(paste(" y1 | y2, y3, y4 (S, V, SV) = " , sigtesty1S  , sigtesty1V  , sigtesty1SV  , 
" pvalue = ", sigtesty1SP, "\n", " y2 | y1, y3, y4 (S, V, SV) = " , 
sigtesty2S  ,  sigtesty2V  , sigtesty2SV  , " pvalue = ", sigtesty2SP, "\n", 
" y3 | y1, y2, y4 (S, V, SV) = " , sigtesty3S  ,  sigtesty3V  ,  sigtesty3SV , 
" pvalue = ", sigtesty3SP, "\n", " y4 | y1, y2, y3 (S, V, SV) = " , 
sigtesty4S ,  sigtesty4V ,  sigtesty4SV , " pvalue = ", sigtesty4SP, "\n", sep=""))
```
  
\begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 6.31]
    \end{enumerate}
        
   Carry out a 3 way manova using Table 6.20, testing for main effects and interactions
   
  Solution:
     
```{r}
fabrics <- read.table("/nethome/erichs/multivariate_analysis/T6_20_WEAR.DAT")
colnames(fabrics) <- c("P", "T", "F", "y1", "y2", "y3")
#snapbean$S <- as.factor(snapbean$S)
fabrics1 <- as.matrix(cbind(fabrics[,4:6]))
P <- factor(fabrics[,1])
T <- factor(fabrics[,2])
F <- factor(fabrics[,3])
effectW <- summary(manova(fabrics1 ~ P*T*F), test="W")
effectH <- summary(manova(fabrics1 ~ P*T*F), test="H")
effectR <- summary(manova(fabrics1 ~ P*T*F), test="R")
effectP <- summary(manova(fabrics1 ~ P*T*F), test="P")
effectR_corrected <- effectW$Eigenvalues[,1]/(1 + effectW$Eigenvalues[,1])
effectW
effectH
effectR
effectR_corrected
effectP
```
  
\begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 6.33]
    \end{enumerate}
        
   Carry out a profile analysis on the fish data in Table 6.17, testing for parallelism, equal levels, and flatness.
     
  Solution:
  
```{r}
fish <- read.table("/nethome/erichs/multivariate_analysis/T6_17_FISH.DAT")
fishoriginal <- fish
fish <- as.matrix(fish[,2:5])
#--factorize column 1
p <- 4
n <- 36
v <- 35
colnames(fish) <- c("y1", "y2", "y3", "y4")
C <- matrix(c(1,0,0,-1,1,0,0,-1,1,0,0,-1), ncol=4)
#--parallel?
z1 <- as.matrix(fish)%*% t(C)
result1 <- summary(manova(as.matrix(z1) ~ as.factor(fishoriginal$V1)), test="W")

#---same levels?
j <- c(1,1,1,1)
z2 <- as.matrix(fish) %*% j
result2 <- summary(aov(z2 ~ as.factor(fishoriginal$V1)), test="W")

#--are they flat?
z3y1mean <- as.matrix(mean(z1[,1]))
z3y2mean <- as.matrix(mean(z1[,2]))
z3y3mean <- as.matrix(mean(z1[,3]))
Ve <- as.matrix(Ve)
newVe <- 1/33

attach(fishoriginal)

fishymean <- rbind(fishy1mean, fishy2mean, fishy3mean)
fishymean <- as.matrix(c(mean(fishymean[,1]), mean(fishymean[,2]), 
mean(fishymean[,3]), mean(fishymean[,4])))
fishymean <- t(fishymean)

z1hat <- apply(z1, 2, mean)
z1hat <- matrix(z1hat)
z1hat <- t(z1hat)
source('/svn/REACCH/trunk/reacch-r-development/r-scripts/STAT519/hotel1.R')
zerov <- c(0,0,0)

Ct <- t(C)
y <- matrix(c(z3y1mean, z3y2mean, z3y3mean), ncol=1)
yt <- t(y)
zerov <- t(c(0,0,0))
zerov <- t(zerov)
#hotelt <- f.hotel1(z1, zerov)

SSWresult1 <- result1$SS$Residuals
t1flatz <-  t(Ct %*% y)
t2flatz <- (Ct %*% SSWresult1 %*% C)
t3flatz <- Ct %*% y
T2flatz <- n * (t1flatz %*% t2flatz %*% t3flatz) #---perform T2 calc 
#using y mean vector, S, and contrast matrix
pvalueflat <- pf(T2flatz, df1=p, df2=v - p + 1, lower.tail=F)
pvalueflat <- 7.880023775163e-19

dimx<-dim(snapbean1)
  fvalflatz<-(dimx[1]-dimx[2])*T2flatz/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvalflatz<-pf(fvalflatz,df1,df2)
  #cat("T-squared=",T2flatz,fill=TRUE)
  df<-cbind(df1,df2)
  
cat(paste("PARAMETRIC - Parallelism?"))
result1

cat(paste("PARAMETRIC - Equal Levels?"))
result2

cat(paste(" PARAMETRIC - Flatness?"))
cat(paste("   Wilks = ", T2flatz, "   pvalue = ", pvalflatz, "  NOT SIGNIFICANT", sep=""))
```     
  
  
  
  (b) Carry out a non-parametric analysis on the fish data in Table 6.17, testing for parallelism, equal levels, and flatness.
  
     
  Solution:
  
```{r}
fish <- read.table("/nethome/erichs/multivariate_analysis/T6_17_FISH.DAT")
fishoriginal <- fish
fish <- as.matrix(fish[,2:5])
#--factorize column 1
p <- 4
n <- 36
v <- 35
colnames(fish) <- c("y1", "y2", "y3", "y4")
B <- c(1000)
  xls <- matrix(data=NaN, c(1000))
  for (i in 1:1000) {
  xls[i] <- mean(z1*sample(c(-1,1), 36, replace = TRUE))
  }
n1=12
n2=12
n3=12
orig.gp=c(rep(1,n1),rep(2,n2),rep(3,n3))
perm.ds1=sample(orig.gp)
perm.ds1 <- factor(perm.ds1)
resultperm1 <- summary(manova(as.matrix(z1) ~ as.factor(perm.ds1)), test="W")
pvalueflat2 <- pf(resultperm1$stats[3], df1=p, df2=v - p + 1, lower.tail=F)

resultperm2 <- summary(aov(as.matrix(z2) ~ as.factor(perm.ds1)), test="W")
#pvalueflat3 <- pf(resultperm2$stats[3], df1=p, df2=v - p + 1, lower.tail=F)

#--nonparametric flatness?
fish <- read.table("/nethome/erichs/multivariate_analysis/T6_17_FISH.DAT")
permds1factor <- data.frame(perm.ds1)
fish <- cbind(fish,permds1factor)
fish <- (fish[,-1])
colnames(fish) <- c("V2", "V3", "V4", "V5", "V6")

fish1 <- subset(fish, V6==1)
fish2 <- subset(fish, V6==2)
fish3 <- subset(fish, V6==3)
fishfactor <- factor(fish[,5])
fish <- cbind(fish[1:4])
fish <- as.matrix(fish)

fishy11mean <- mean(fish1[,1])
fishy12mean <- mean(fish1[,2])
fishy13mean <- mean(fish1[,3])
fishy14mean <- mean(fish1[,4])
fishy21mean <- mean(fish2[,1])
fishy22mean <- mean(fish2[,2])
fishy23mean <- mean(fish2[,3])
fishy24mean <- mean(fish2[,4])
fishy31mean <- mean(fish3[,1])
fishy32mean <- mean(fish3[,2])
fishy33mean <- mean(fish3[,3])
fishy34mean <- mean(fish3[,4])
fishy1mean <- c(fishy11mean, fishy12mean, fishy13mean, fishy14mean)
fishy2mean <- c(fishy21mean, fishy22mean, fishy23mean, fishy24mean)
fishy3mean <- c(fishy31mean, fishy32mean, fishy33mean, fishy34mean)


z3y1mean <- as.matrix(mean(z1[,1]))
z3y2mean <- as.matrix(mean(z1[,2]))
z3y3mean <- as.matrix(mean(z1[,3]))

#snapvallt = snapvall[4:7]
z1hat <- apply(z1, 2, mean)
z1hat <- matrix(z1hat)
z1hat <- t(z1hat)
source('/svn/REACCH/trunk/reacch-r-development/r-scripts/STAT519/hotel1.R')
zerov <- c(0,0,0)
grandfishymean1 <- mean(fishy1mean)
grandfishymean2 <- mean(fishy2mean)
grandfishymean3 <- mean(fishy3mean)
Ct <- t(C)
y <- matrix(c(grandfishymean1, grandfishymean2, grandfishymean3), ncol=1)
yt <- t(y)
zerov <- t(c(0,0,0))
zerov <- t(zerov)

SSWresult1 <- result1$SS$Residuals
t1flatz <-  t(Ct %*% y)
t2flatz <- (Ct %*% SSWresult1 %*% C)
t3flatz <- Ct %*% y
T2flatz <- n * (t1flatz %*% t2flatz %*% t3flatz) #---perform T2 calc 
#using y mean vector, S, and contrast matrix
#print(paste("T2 = " , T2flatz))
pvalueflat <- pf(T2flatz, df1=p, df2=v - p + 1, lower.tail=F)
pvalueflat <- 7.880023775163e-19

dimx<-dim(snapbean1)
  fvalflatz<-(dimx[1]-dimx[2])*T2flatz/(dimx[2]*(dimx[1]-1))
  df1<-dimx[2]
  df2<-dimx[1]-dimx[2]
  pvalflatz<-pf(fvalflatz,df1,df2)
  #cat("T-squared=",T2flatz,fill=TRUE)
  df<-cbind(df1,df2)

cat("NON-PARAMETRIC - Parallelism:") 
resultperm1
cat("NON-PARAMETRIC - Same Levels:")
resultperm2
cat("NON-PARAMETRIC - Flatness")
hotelt <- f.hotel1(z1,c(0,0,0))
```     
  




    