---
title: "University of Idaho Statistics 519 Spring 2015 Homework #1"
author: "Author: David Erich Seamon - erichs@uidaho.edu "
date: "Due Date: 1/30/2015"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{enumitem}
- \pagestyle{fancy}
- \fancyhead[CO,CE]{Erich Seamon - University of Idaho}
- \fancyfoot[CO,CE]{Statistics 519 - Homework 1 - 01/30/2015}
- \renewcommand{\headrulewidth}{0.4pt}% Default \headrulewidth is 0.4pt
- \renewcommand{\footrulewidth}{0.4pt}% Default \footrulewidth is 0pt
- \fancyfoot[LE,RO]{\thepage}
output:
  pdf_document:
    keep_tex: yes
---

\noindent {\bf ASSIGNED PROBLEMS: 2.6, 2.12, 2.19, 2.20, 2.25, 2.26, 2.31, 2.34, 2.36, 2.38, 3.5, 3.9, 3.14, 3.20, 3.21}

\begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}
    \item [Problem 2.6]
    
    \end{enumerate}
    \begin{enumerate}
    

    \item[] Let
      

  
        \[ A = \begin{pmatrix} 1 & 2 & 3 \\
                               2 & 4 & 6 \\
                               5 & 10 & 15 \end{pmatrix},
          \quad B = \begin{pmatrix} -1 & 1 & -2\\ 
    		     -1 & 1 & -2 \\
 				      1 & -1 & 2 \end{pmatrix} \]
              
\setlength{\itemindent}{.5in}

   \item[(a)] Show that AB = O
   \item []
   Solution:
        \[ A = \begin{pmatrix} (1)(-1) & + & (2)(-1) &+ & (3)(1) \\
                               (2)(1) & + & (4)(1) & + & (6)(-1) \\
                               (5)(-2) & + & (10)(-2) & + & (15)(2) \end{pmatrix}\]

        \[ A = \begin{pmatrix} -1 & + & -2 & + & 3 \\
                               -2 & + & -4 & + & 6 \\
                               -10 & + & -20 & + & 30 \end{pmatrix}\]

        \[ A = \begin{pmatrix} 0 \\
                               0 \\
                               0 \end{pmatrix}\]
                               
\setlength{\itemindent}{.5in}

   \item[(b)] Find a vector x such that Ax = 0
   \item []
Solution: Given 2.6(a), we know that A is linearly dependent, since column 3 is a product of columns 1 and 2, and the det(A) = 0.  Under the concept of linear dependence, there exists a constant for each column such that, when multiplied by each column, produces 0.\

        \[c^{1}\begin{pmatrix} 1 \\
                               2 \\
                               5 \end{pmatrix} + 
    \ c^{2} \begin{pmatrix} 2 \\ 4 \\ 10 \end{pmatrix} +
 \ c^{3} \begin{pmatrix} 3 \\ 6 \\ 15 \end{pmatrix} = 0\]
 
 
Since Matrix A is homogeneous and singular, as well as non-invertible, there are many solutions to vector x.  Because column 3 is the summation of columns 1 and 2, all solutions for vector(x) are a vector combination (x, x, -x).

        \[\begin{pmatrix} c^{1}\\
                               c^{2} \\
                               c^{3} \end{pmatrix} = 
        \ \begin{pmatrix} x\\
                                x \\
                               -x \end{pmatrix} = (x, x, -x)'\;\;(i.e.\;\;(1, 1, -1),\;(2, 2, -2)...)\]
\\
\\
\\

  \setlength{\itemindent}{.5in}



   \item[(c)] Show that |A| = 0
   \item[]
   Solution:
   
   \end{enumerate}
   
```{r}
A <- matrix(c(1,2,5,2,4,10,3,6,15), ncol=3)
det(A)
```
   
   \begin{enumerate}\bfseries
  
        
  \setlength{\itemindent}{.5in}
                
  \item [Problem 2.12]
    \end{enumerate}
    \begin{enumerate}
    \item[] Let
    
    \[ A = \begin{pmatrix} 1 & 2 & 3 \\
                               4 & 5 & 6 \\
                               7 & 8 & 9 \end{pmatrix},
          \quad B = \begin{pmatrix} a & 0 & 0\\ 
             0 & b & 0 \\
 				      0 & 0 & c \end{pmatrix} \]
                 
    \setlength{\itemindent}{.5in}
    \mbox{}\\ Find DA, AD, and DAD
    
    \setlength{\itemindent}{.5in}

     \item[]
     Solution:

    \item[] Find DA
    \item [] By using matrix multiplication of rows to columns, we can determine DA.
    
            \[ DA = \begin{pmatrix} (a + 0 + 0) &  & (2a + 0 + 0)&  & (3a + 0 + 0) \\
                               (0 + 4b + 0) &  & (0 + 5b + 0) &  & (0 + 6b + 0) \\
                               (0 + 0 + 7c) &  & (0 + 0 + 8b) &  & (0 + 0 + 9c) \end{pmatrix}\]

        \[ DA = \begin{pmatrix} a & 2a & 3a \\
                               4b & 5b & 6b \\
                               7c & 8c & 9c \end{pmatrix}\]
        
      \setlength{\itemindent}{.5in}

     \item[]
     Solution:

    \item[] Find AD
    \item [] By using matrix multiplication of rows to columns, we can determine AD.
    
            \[ AD = \begin{pmatrix} (a + 0 + 0) &  & (0 + 2b + 0)&  & (0 + 0 + 3c) \\
                              (4a + 0 + 0) &  & (0 + 5b + 0) &  & (0 + 0 + 6c) \\
                              (7a + 0 + 0) &  & (0 + 8b + 0) &  & (0 + 0 + 9c) \end{pmatrix}\]

        \[ AD = \begin{pmatrix} a & 2b & 3c \\
                               4a & 5b & 6c \\
                               7a & 8b & 9c \end{pmatrix}\]    
        
        \setlength{\itemindent}{.5in}

     \item[]
     Solution:

    \item[] Find DAD
    \setlength{\itemindent}{.5in}
    \item [] By using matrix multiplication of either (DA)D or D(AD), we can determine DAD.  Below, we calculate (DA)D.
    
            \[ DA = \begin{pmatrix} a & 2a & 3a \\
                               4b & 5b & 6b \\
                               7c & 8c & 9c \end{pmatrix},
          \quad B = \begin{pmatrix} a & 0 & 0\\ 
      	     0 & b & 0 \\
 				      0 & 0 & c \end{pmatrix} \]
    
          \[ DAD = \begin{pmatrix} (a^{2} + 0 + 0) &  & (0 + 2ab + 0)&  & (0 + 0 + 3ac) \\
                              (4ab + 0 + 0) &  & (0 + 5b + 0) &  & (0 + 0 + 6bc) \\
                              (7ac + 0 + 0) &  & (0 + 8bc + 0) &  & (0 + 0 + 9c^{2}) \end{pmatrix}\]

        \[ DAD = \begin{pmatrix} a^{2} & 2ab & 3ac \\
                               4ab & 5b^{2} & 6bc \\
                               7c & 8bc & 9c^{2} \end{pmatrix}\]


          
            \end{enumerate}

            \begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}

    \item [Problem 2.15]
    \end{enumerate}
    \begin{enumerate}
    \item[] NOTE: certain solutions to problem 2.15 that use matrix A, as published in Rencher (2002) are not correct, given the two matrices provided. In particular, given the solutions as published, it appears that matrix A should have 
    
    \[a_{33} = 3\;\;\;NOT\;\;\;a_{33} = 2\]
    
    \item[] 
    
    
    
    
    \item[] Let
    
    \[ A = \begin{pmatrix} 5 & 4 & 4 \\
                               2 & -3 & 1 \\
                               3 & 7 & 2 \end{pmatrix},
          \quad B = \begin{pmatrix} 1 & 0 & 1\\ 
      	     0 & 1 & 0 \\
 				      1 & 2 & 3 \end{pmatrix} \]
    
    
    \setlength{\itemindent}{.5in}

    \item[(a)] Find tr(A) and tr(B)
   \item[]
   Solution:
   
   \item[] tr(A) = 5 + (-3) + 2 = 4\;\;\;\;NOTE: Listed as 5 but actually 4 given Matrix A error\\
  \item[] tr(B) = 1 + 1 + 3 = 5
   
    \end{enumerate}

```{r}
A <- matrix(c(5, 2, 3, 4, -3, 7, 4, 1, 2), ncol=3)
B <- matrix(c(1, 0, 1, 0, 1, 2, 1, 0, 3), ncol=3)
sum(diag(A))
sum(diag(B))
```
 
 \begin{enumerate}

    
    \setlength{\itemindent}{.5in}

    \item[(b)] Find A + B and tr(A + B).  Is tr(A + B) = tr(A) + tr(B)?
   \item[]
   Solution: A + B

\[ A + B = \begin{pmatrix} (5 + 1) & (4 + 0) & (4 + 1) \\
                               (2 + 0) & (-3 + 1) & (1 + 0) \\
                               (3 + 1) & (7 + 2) & (2 + 3) \end{pmatrix}\]

        \[ A + B = \begin{pmatrix} 6 & 4 & 5 \\
                               2 & -2 & 1 \\
                               4 & 9 & 5 \end{pmatrix}\]

    \item[]
    \item[] Solution: tr(A + B) \\
    \item[] tr(A + B) = (6 + (-2) + 5) =9\;\;\;\;NOTE: Listed as 10 but actually 9 given Matrix A error
    
    \item[]
    \item[] Solution: Is tr(A + B) = tr(A) + tr(B)? \\
    \item[] Yes.  tr(A) = 4, tr(B) = 5, and tr(A + B) = 9, as noted above.\\

    \end{enumerate}

```{r}
SumAB <- A + B
tr1 <- sum(diag(SumAB))
tr2 <- sum(diag(A)) + sum(diag(B))
tr1
tr2
```
 
 \begin{enumerate}
\setlength{\itemindent}{.5in}

    \item[(c)] Find |A| and |B|
   \item[]
   Solution: |A|

  \item[] |A| = (5((-3)(2) - (7)(1))) - (4((2)(2) - (3)(1))) + (4((2)(7) - (2)(7)))
  \item[] |A| = 5(-13) - 4(1) + 4(23)
  \item[] |A| = 23\;\;\;\;NOTE:\;Listed\;as\;0\;but\;actually\;23\;given\;Matrix\;A\;error\;
  
  \item[]
   Solution: |B|

  \item[] |B| = (1((1)(3) - (2)(0))) - (0((3)(0) - (1)(0))) + (1((2)(0) - (1)(1)))
  \item[] |B| = 1(3) - 0(0) + 1(-1)
  \item[] |B| = 2
  



 \end{enumerate}
   
```{r}
A <- matrix(c(5,2,3,4,-3,7,4,1,2), ncol=3)
B <- matrix(c(1,0,1,0,1,2,1,0,3), ncol=3)
det(A)
det(B)
```
   
   \begin{enumerate}

\setlength{\itemindent}{.5in}

    \item[(d)] Find AB and |AB|.  Is |AB| = |A||B|?
   \item[]
   Solution: AB
   
   \[ AB = \begin{pmatrix} (5 + 0 + 4) &  & (0 + 4 + 8)&  & (5 + 0 + 12) \\
                               (2 + 0 + 1) &  & (0 + -3 + 2) &  & (2 + 0 + 3) \\
                               (3 + 0 + 2) &  & (0 + 7 + 4) &  & (3 + 0 + 6) \end{pmatrix}\]

        \[ AB = \begin{pmatrix} 9 &  & 12 &  & 17 \\
                               3 &  & -1 & + & 5 \\
                               5 &  & 11 & + & 9 \end{pmatrix}\]
                               
                               \end{enumerate}
   
```{r}
A <- matrix(c(5,2,3,4,-3,7,4,1,2), ncol=3)
B <- matrix(c(1,0,1,0,1,2,1,0,3), ncol=3)
AB <- A %*% B
AB
```
   
   \begin{enumerate}

\setlength{\itemindent}{.5in}
                               
    \item[]
   Solution: |AB|
   
   |AB| = 9((9)(-1) - (11)(5)) - 12((9)(3) - (5)(5)) + 17((3)(11) - (5)(-1)) \\
   |AB| = 9(-9 - 55) - 12(27 - 25) + 17(33 - 5) \\
   |AB| = 46
   
   \end{enumerate}
   
```{r}
A <- matrix(c(5,2,3,4,-3,7,4,1,2), ncol=3)
B <- matrix(c(1,0,1,0,1,2,1,0,3), ncol=3)
AB <- A %*% B
det(AB)
```
   
   \begin{enumerate}

\setlength{\itemindent}{.5in}

                               
    \item[]
   Solution: Is |AB| = |A||B|?  Answer: Yes.  See calculations below.
      
      \end{enumerate}

```{r}
A <- matrix(c(5,2,3,4,-3,7,4,1,2), ncol=3)
B <- matrix(c(1,0,1,0,1,2,1,0,3), ncol=3)
AB <- A %*% B
det(AB)
detAB <- det(A) %*% det(B)
detAB
```
   
   \begin{enumerate}\bfseries

\setlength{\itemindent}{.5in}
   
        
\setlength{\itemindent}{.5in}

    \item [Problem 2.19]
            \end{enumerate}

        \begin{enumerate}

    \item[] Let
    
    \[ A = \begin{pmatrix} 1 & 1 & -2 \\
                               -1 & 2 & 1 \\
                               0 & 1 & -1 \end{pmatrix}\]
                               
    \item[(a)] Find the eigenvalues and associated normalized eigenvectors.
   \item[]
   Solution:
   
   Use the characteristic equation:  \[|A - \lambda I| = 0\]
        
    \item[] 
    
     \[ |A - \lambda I| = \begin{pmatrix} 1 - \lambda & 1 & -2 \\
                               -1 & 2 - \lambda & 1 \\
                               0 & 1 & -1 - \lambda \end{pmatrix} = 0\]
    
          \end{enumerate}

```{r}
A <- matrix(c(1,-1,0,1,2,1,-2,1,-1), ncol=3)
eigenA <- eigen(A)
eigenvaluesA <- eigenA$values
eigenvectorsA <- eigenA$vectors
eigenvaluesA
round(eigenvectorsA, digits=4)
```
   
   \setlength{\itemindent}{.5in}

   \begin{enumerate}

    \item[(b)] Find tr(A) and |A| and show that: \[tr(A) = \sum_{n=1}^{10} \lambda_{i} \;and \;|A| = \prod_{n=1}^{3} \lambda_{i}\]
   \item[]
   Solution: Find tr(A) and |A|
   
   tr(A) = (1 + 2 + -1) = 2
   \end{enumerate}

```{r}
A <- matrix(c(1,-1,0,1,2,1,-2,1,-1), ncol=3)
sum(diag(A))
det(A)
```

   \begin{enumerate}
        
   \item[]
   Solution: show that: \[tr(A) = \sum_{n=1}^{10} \lambda_{i} \;and \;|A| = \prod_{n=1}^{3} \lambda_{i}\]
   \\
   
   \item[] \[tr(A) = 2 \;(see\;calculation\;above)\]
   \item[] \[\sum_{n=1}^{10} \lambda_{i} = (2, 1, -1) = 2\]
   \\
   
   \item[] \[det(A) = -2 \;(see\;calculation\;above)\]
   
   \[\prod_{n=1}^{3} \lambda_{i} = (2)(1)(-1) = -2\]\\
           \end{enumerate}

             \begin{enumerate}\bfseries
          \setlength{\itemindent}{.5in}
          
\setlength{\itemindent}{.5in}

    \item [Problem 2.20]
        \end{enumerate}
            \begin{enumerate}

    \item[] Let
    
    \[ A = \begin{pmatrix} 3 & 1 & 1 \\
                               1 & 0 & 2 \\
                               1 & 2 & 0 \end{pmatrix}\]
                               
    \item[(a)] The eigenvalues of A are 1, 4, -2.  Find the normalized eigenvectors and use them as columns in an orthogonal matrix C.
    \\
    
    NOTE: The calculated eigenvalues of A are 4, 1, -2.  As such, this alters the solution for the normalized eigenvectors in (a). 
    
   \item[]
   Solution:  
   
    \end{enumerate}

```{r}
A <- matrix(c(3,1,1,1,0,2,1,2,0), ncol=3)
eigenA <- eigen(A)
eigenvectorsA <- eigenA$vectors
round(eigenvectorsA, digits=4)
```
      

   \begin{enumerate}
   
   \setlength{\itemindent}{.5in}
   
   \item[(b)] Show that C'AC = D as in (2.111), where D is diagonal with the eigenvalues of A on the diagonal.
    
   \item[]
   Solution:  If C is a matrix of normalized eigenvectors of A - then C is orthogonal, and C'C = I.  
   
   \end{enumerate}

```{r}
A <- matrix(c(3,1,1,1,0,2,1,2,0), ncol=3)
eigenA <- eigen(A)
eigenvectorsA <- eigenA$vectors
teigenvectorsA <- t(eigenvectorsA)
comparedD <- teigenvectorsA %*% A %*% eigenvectorsA
round(comparedD, digits = 1)
comparedD
```
    
   \begin{enumerate}
   
        \setlength{\itemindent}{.5in}
   
   \item[(c)] Show that A = CDC' as in (2.109)
    
   \item[]
   Solution:  If C is a matrix of normalized eigenvectors of A - then C is orthogonal, and C'C = I.  
   
   \end{enumerate}

```{r}
A <- matrix(c(3,1,1,1,0,2,1,2,0), ncol=3)
eigenA <- eigen(A)
eigenvectorsA <- eigenA$vectors
teigenvectorsA <- t(eigenvectorsA)
D <- teigenvectorsA %*% A %*% eigenvectorsA
round(D, digits = 1)
comparedA = eigenvectorsA %*% D %*% teigenvectorsA
round(comparedA, digits=1)
```


   \begin{enumerate}\bfseries
   
\item[ ]
\item[ ]
        
\setlength{\itemindent}{.5in}

    \item [Problem 2.25]
        \end{enumerate}
       \begin{enumerate}

   \item[] Verify (2.41); that is, show that (x - y)' (x - y) = x'x - 2x'y + y'y
    
   Solution:  
   \\
   
   Several facts enable 
   
   \[(x - y)' (x - y) = x'x - 2x'y + y'y\]
   
   -The transpose of the sum of two matrices equals the sum of the transposes.\\
   \\
   -You can expand the products of (x - y)' (x - y) to a structure of:
   
   \[(x - y)' (x - y) = ((x' - y')x - (x' - y')y)\]
   \[x'x - y'x - x'y + y'y\]
   \[x'x - 2x'y + y'y\]
   
   \end{enumerate}
   
   
   
   \begin{enumerate}\bfseries
   
        
        
\setlength{\itemindent}{.5in}

    \item [Problem 2.26]
     \end{enumerate}
       \begin{enumerate}
   \item[] Show that A'A is symmetric, where A is n x p.
    
   Solution:  
   \\
   
   Several facts show that A'A is symmetric.\\
  \\
   -a symmetric matrix is a square matrix that is equal to its transpose.\\
   \\
   -the statement below shows that taking the transpose of A'A results in A'A.\\
   \\
   \[(A'A)' = (A')' A' = A'A\]
   
   \end{enumerate}
   
   \begin{enumerate}\bfseries
   
        
\setlength{\itemindent}{.5in}

    \item [Problem 2.31]
     \end{enumerate}
       \begin{enumerate}
   \item[] Show that the inverse of a partitioned matrix given in (2.76) is correct by multiplying by
   
   \[ \begin{pmatrix} A_{11} & a_{12} \\
                               a'_{12} & a_{22} \end{pmatrix}\]
                               \\
                               
   \item[] to obtain an identity matrix.\\
   \\
   Solution:  
   \\
   
   \[ A' = 1/b \begin{pmatrix} bA_{11}^{-1} + A_{11}^{-1} a_{12} A_{11}^{-1} & A_{11}^{-1} a_{12} \\
    -a'_{12} A_{11}^{-1} & 1 \end{pmatrix}
    \ \begin{pmatrix} A_{11} & a_{12} \\
    -a'_{12} & a_{22} \end{pmatrix}\]
    
    \[ = 1/b \begin{pmatrix} bI + A_{11}^{-1} a_{12} - A_{11}^{-1} a_{12} a'_{12} & bA_{11}^{-1} a_{12} + A_{11}^{-1} a_{12} a'_{12} A_{11}^{-1}  a_{12} - A_{11}^{-1} a_{12} a_{22} \\
    -a'_{12} + a'_{12} & -a'_{12} A_{11}^{-1} a_{12} + a_{22} \end{pmatrix}\]
    
            \[ = 1/b \begin{pmatrix} bI & 0 \\
    0' & b \end{pmatrix}\]    
    
    \[ = \begin{pmatrix} I & 0 \\
    0' & 1 \end{pmatrix}\]   
   

   \end{enumerate}
   
      \begin{enumerate}\bfseries

   
        \setlength{\itemindent}{.5in}

    \item [Problem 2.34]
     \end{enumerate}
       \begin{enumerate}
   \item[] Show that\[|A^{-1}| = 1/|A| \]
    
   Solution:  
              
   \item[] The determinant of the transpose of a matrix is the same as the determinant of the matrix, and the determinant of the inverse of a matrix is the reciprocal of the determinant:
   
   \[|A'| = |A|\]
   \[|A^{-1}| = 1/|A| = |A|^{-1}\]
   
                 \end{enumerate}
          
\begin{enumerate}\bfseries
  
        
\setlength{\itemindent}{.5in}

    \item [Problem 2.36]
     \end{enumerate}
       \begin{enumerate}
   \item[] Show that \[tr(A'A) = tr(AA)' = \sum_{ij}^n a_{ij}^{2}\]
    
   Solution:  
   \\
   
   \[tr(AA') = \sum_{i=1}^n a'_{i} a_{i} = \sum_{i=1}^n (a_{i1}^{2} + a_{i2}^{2} + ... + a_{in}^{2})\]
   
   \[= \sum_{ij}^n a_{ij}^{2}\]

\end{enumerate}

```{r}
A <- matrix(c(3,1,1,1,0,2,1,2,0), ncol=3)
tA <- t(A)
sum(diag(tA %*% A))
sum(diag(A %*% tA))
```
   

\begin{enumerate}\bfseries
   
        
\setlength{\itemindent}{.5in}

    \item [Problem 2.38]
      \end{enumerate}
       \begin{enumerate}
    \item[] Show that the eigenvalues of AB are the same as those of BA, as noted in Section 2.11.5.
    \\
    
   \item[] Solution:  Suppose lambda is an eigenvalue of AB. Then 
   
   \[ABx = \lambda x\;and\;BABx = \lambda Bx\] resulting in Bx
being an eigenvector of BA with eigenvalue lambda.  AB and BA have the
same eigenvalues, so their traces are equal - which are the sum of their eigenvalues.

        \end{enumerate}
        
        \begin{enumerate}\bfseries
   
        
\setlength{\itemindent}{.5in}

    \item [Problem 3.5]
      \end{enumerate}
       \begin{enumerate}
    \item[] For p = 3, show that
    
    \[1/(n - 1) \sum_{i-1}^{n} (y_{i} - \bar{y})(y_{i} - \bar{y})' = \begin{pmatrix} s_{11} & s_{12} & s_{13} \\
                               s_{21} & s_{22} & s_{23} \\
                               s_{31} & s_{32} & s_{33} \end{pmatrix}\]
    \\
    
   \item[] Solution:  If p = 3, then the sum of squares calculation for the sample variance can be expanded into two matrices with a p = 3.

     \[ \sum_{i=1}^{n} \begin{pmatrix} y_{i1} \\
                               y_{i2} \\
                               y_{i3} \end{pmatrix}
          \begin{pmatrix} y_{i1} - \bar{y}_{1}, y_{i2} - \bar{y}_{2}, y_{i3} - \bar{y}_{3}\end{pmatrix}\]
          
          
          \[ = \sum_{i=1}^{n} \begin{pmatrix} (y_{i1} - \bar{y}_{1})^{2} & (y_{i1} - \bar{y}_{1})(y_{i2} - \bar{y}_{2}) & (y_{i1} - \bar{y}_{1})(y_{i3} - \bar{y}_{3}) \\
                               (y_{i2} - \bar{y}_{2})(y_{i1} - \bar{y}_{1}) & (y_{i2} - \bar{y}_{2})^{2} & (y_{i2} - \bar{y}_{2})(y_{i3} - \bar{y}_{3})\\
                               (y_{i3} - \bar{y}_{3})(y_{i1} - \bar{y}_{1}) & (y_{i3} - \bar{y}_{3})(y_{i2} - \bar{y}_{2}) & (y_{i3} - \bar{y}_{3})^{2} \end{pmatrix}\]
                               
        \end{enumerate}

        \begin{enumerate}\bfseries
   
        
\setlength{\itemindent}{.5in}

    \item [Problem 3.9]
      \end{enumerate}
       \begin{enumerate}
    \item[] Verify
    
    \[cov(z) = I/n, where\;z = (\Sigma^{1/2})^{-1} (\bar{y} - \mu))\]
    
    Solution:
    \\
    
    substituting z 
    \\
    \[cov(z) = cov[(\Sigma^{1/2})^{-1} \bar{y} - (\Sigma^{1/2})^{-1} \mu) \]
    
    transposing (y - mean) to /n
    \\
    \[cov(z) = (\Sigma^{1/2})^{-1} \begin{pmatrix} \Sigma \\  n  \end{pmatrix} (\Sigma^{1/2})^{-1}\] 
    
    \[cov(z) = [(\Sigma^{1/2})^{-1} \begin{pmatrix} \Sigma \\  n  \end{pmatrix}\] 
    
    Then pulling out n
    \\
    \[cov(z) = 1/n(\Sigma^{1/2})^{-1} \Sigma^{1/2} \Sigma^{1/2} (\Sigma^{1/2})^{-1}\] 
    
    and cancelling out all the square root matrices, leaves
    
    \[cov(z) = (1/n)I\]
    
    \end{enumerate}
    
    \begin{enumerate}\bfseries
   
        
\setlength{\itemindent}{.5in}

    \item [Problem 3.14]
      \end{enumerate}
       \begin{enumerate}
    \item[] For the variables in Table 3.3, define z as:
    
    \[z = 3y_{1} - y_{2} + 2y_{3}\]
    \\
        \[z = (3, -1, 2)y\]

   \item [] (a) Evaluate z for each row of Table 3.3 and find
     
     \[\bar{z}\]
     
     and 
     
     \[s^{2}_{z}\] 
     
     directly from 
     
     \[z_{1}, z_{2} ... z_{10}\]
     
     \item[] using (3.1) and (3.5).
     \\
     Solution:
     
     \end{enumerate}
     
```{r}
Table3 <- matrix(c(35,35,40,10,6,20,35,35,35,30,3.5,4.9,30.0, 2.8, 
2.7, 2.8, 4.6, 10.9, 8.0, 1.6, 2.80, 2.70, 4.38, 3.21, 2.73, 2.81, 
2.88, 2.90, 3.28, 3.20), ncol = 3)

z <- numeric(0)
z <- c(z, 1:10)
obs <- c(1:10)
vars <- c(1:3)
for (i in obs) {
  for (j in vars) {
  z[i] <- ((Table3[i,1]) * 3) - (Table3[i,2]) + ((Table3[i,3]) * 2)
}}
mean(z)
var(z)
```

\begin{enumerate}

   \item[] (b) Calculate the sample mean of z by using
     
     \[\bar{z} = a'\bar{y},\;and\;s_{z}^{2} = a'Sa\]
     
     \[\bar{z} = a'\bar{y} = \begin{pmatrix} 3 & 1 & 2 \end{pmatrix} \begin{pmatrix} 28.1 \\ 7.18 \\ 3.089 \end{pmatrix}\] 
     
    
     
  \end{enumerate}
     
```{r}
a <- matrix(c(3,-1,2))
ta <- t(a)
meany <- matrix(c(28.1, 7.180, 3.089))
ta %*% meany

S <- cov(Table3)
z2 <- ta %*% S %*% a
z2
```
 
 
        
 \setlength{\itemindent}{.5in}

   \begin{enumerate}\bfseries
\setlength{\itemindent}{.5in}

   \item[Problem 3.20]
   
     \end{enumerate}
       \begin{enumerate}
     \item[] For the data Table in 3.6, define
     
     \[z_{1} = 2_{y1} + 3_{y2} - y_{3} + 4y_{4},\]
     
     \[z_{2} = -2_{y1} - {y2} + 4y_{3} - 2y_{4},\]
     
     \[z_{3} = 3_{y1} - 2_{y2} - y_{3} + 3y_{4}.\]
     
     \item[] Find
     
     \[\bar{z}, S_{z},\;and\;R_{z}\]
     
     Solution:
     
     \[Solve for \bar{z} = A'\bar{y} = \begin{pmatrix} 2 & 3 & -1 & 4 \\ -2 & -1 & 4 & -2 \\ 3 & -2 & -1 & 3 \end{pmatrix} \begin{pmatrix} y1 \\ y2 \\ y3 \\ y4 \\ y5 \end{pmatrix}\]
     
     \end{enumerate}
     
```{r}
Table36 <- matrix(c(47.8, 46.4, 46.3, 45.1, 47.6, 52.5, 51.2, 49.8, 48.1, 45.0, 51.2, 
48.5, 52.1, 48.2, 49.6, 50.7, 47.2, 53.3, 46.2, 46.3, 48.8, 47.3, 46.8, 45.3, 48.5, 
53.2, 53.0, 50.0, 50.8, 47.0, 51.4, 49.2, 52.8, 48.9, 50.4, 51.7, 47.7, 54.6, 47.5, 
47.6, 49.0, 47.7, 47.8, 46.1, 48.9, 53.3, 54.3, 50.3, 52.3, 47.3, 51.6, 53.0, 53.7, 
49.3, 51.2, 52.7, 48.4, 55.1, 48.1, 51.3, 49.7, 48.4, 48.5, 47.2, 49.3, 53.7, 54.5, 
52.7, 54.4, 48.3, 51.9, 55.5, 55.0, 49.8, 51.8, 53.3, 49.5, 55.3, 48.4, 51.8), ncol = 4)

mean1 <- mean(Table36[,1])
mean2 <- mean(Table36[,2])
mean3 <- mean(Table36[,3])
mean4 <- mean(Table36[,4])

meany <- matrix(c(mean1, mean2, mean3, mean4))
meany
```
        
 \setlength{\itemindent}{.5in}

   
\setlength{\itemindent}{.5in}
   \begin{enumerate}

\item[] Solve for:

     \[\bar{z} = A'\bar{y} = \begin{pmatrix} 2 & 3 & -1 & 4 \\ -2 & -1 & 4 & -2 \\ 3 & -2 & -1 & 3 \end{pmatrix} \begin{pmatrix} 48.655 \\ 49.625 \\ 50.570 \\ 51.450 \end{pmatrix}\]
     
          \end{enumerate}

```{r}
A <- matrix(c(2, -2, 3, 3, -1, -2, -1, 4, -1, 4, -2, 3), ncol = 4)
bary <- matrix(c(48.655, 49.625, 50.570, 51.450), ncol = 1)
z <- A %*% bary
z
```
        
 \setlength{\itemindent}{.5in}

\setlength{\itemindent}{.5in}

   \begin{enumerate}

     \item[] Solve for:
     
     \[ S_{z} = ASA'\]
     
               \end{enumerate}

```{r}
A <- matrix(c(2, -2, 3, 3, -1, -2, -1, 4, -1, 4, -2, 3), ncol = 4)
S <- cov(Table36)
tA <- t(A)
Sz <- A %*% S %*% tA
Sz
```
        
 \setlength{\itemindent}{.5in}
 
  \setlength{\itemindent}{.5in}

\setlength{\itemindent}{.5in}

   \begin{enumerate}

     \item[] Solve for:
     
     \[ R_{z} = D_{z}^{-1} S_{z} D_{z}^-1\]
     
     \[D_{z}\;is\;obtained\;from\;the\;square\;roots\;of\;diagonal\;elements\;of\;S_{z}\]
     
               \end{enumerate}

```{r}
A <- matrix(c(2, -2, 3, 3, -1, -2, -1, 4, -1, 4, -2, 3), ncol = 4)
S <- cov(Table36)
tA <- t(A)
Sz <- A %*% S %*% tA
Sz
```
        
 \setlength{\itemindent}{.5in}
     \setlength{\itemindent}{.5in}

   \begin{enumerate}

     \item[] Solve for Rz:
     
     
 \end{enumerate}

     
```{r}
D <- diag(Sz)
D <- matrix(D)
Dz1 <- sqrt(D[1,1]) 
Dz2 <- sqrt(D[2,1]) 
Dz3 <- sqrt(D[3,1])
Dz <- matrix(c(Dz1, 0, 0, 0, Dz2, 0, 0, 0, Dz3), ncol = 3)
inverseDz <- solve(Dz)
Rz <- inverseDz %*% Sz %*% inverseDz
Rz
```



\setlength{\itemindent}{.5in}

   \begin{enumerate}\bfseries
\setlength{\itemindent}{.5in}

     \item [Problem 3.21]
     
       \end{enumerate}
       \begin{enumerate}
     \item[] The data in Table 3.7 consist of head measurements on first and second sons (Frets 1921).  Define y1 and y2 as the measurements on the first son and x1 and x2 for the second son.
     
     \item[(a)]
     
     \item Find the mean vector for all four variables and partition it into 
     
     \[\begin{pmatrix} \bar{y} \\
    \bar{x} \end{pmatrix}\]
    
    Solution:
     
      \end{enumerate}
```{r}
Table37 <- matrix(c(191, 195, 181, 183, 176, 208, 189, 197, 188, 192, 179, 183, 174, 190, 
                    188, 163, 195, 186,181, 175, 192, 174, 176, 197, 190, 155, 149, 148, 
                    153, 144, 157, 150, 159, 152, 150, 158, 147, 150, 159, 151, 137, 155, 
                    153, 145, 140, 154, 143, 139, 167, 163, 179, 201, 185, 188, 171, 192, 
                    190, 189, 197, 187, 186, 174, 185, 195, 187, 161, 183, 173, 182, 165, 
                    185, 178, 176, 200, 187, 145, 152, 149, 149, 142, 152, 149, 152, 159, 
                    151, 148, 147, 152, 157, 158, 130, 158, 148, 146, 137, 152, 147, 143, 
                    158, 150), ncol = 4)
mean1 <- mean(Table37[,1])
mean2 <- mean(Table37[,2])
mean3 <- mean(Table37[,3])
mean4 <- mean(Table37[,4])
mean1
mean2
mean3
mean4
```

\setlength{\itemindent}{.5in}

   \begin{enumerate}
\setlength{\itemindent}{.5in}

 \item[] Given the mean calculations above:
     
  \[\begin{pmatrix} \bar{y} \\ \bar{x} \end{pmatrix} = \begin{pmatrix} 185.72 \\ 151.12 \\ \hline 183.84 \\ 149.24 \end{pmatrix}\]
   \end{enumerate}
   
      \begin{enumerate}
      
      \setlength{\itemindent}{.5in}


   \item [(b)]
     
     \item Find the covariance matrix for all four variables and partition it into
     
     \[S = \begin{pmatrix} s_{yy} & S_{yx} \\ S_{xy} & S_{xx} \end{pmatrix} \]
     
     Solution:
     
      \end{enumerate}

```{r}
Table37 <- matrix(c(191, 195, 181, 183, 176, 208, 189, 197, 188, 192, 179, 183, 174, 190, 
                    188, 163, 195, 186,181, 175, 192, 174, 176, 197, 190, 155, 149, 148, 
                    153, 144, 157, 150, 159, 152, 150, 158, 147, 150, 159, 151, 137, 155, 
                    153, 145, 140, 154, 143, 139, 167, 163, 179, 201, 185, 188, 171, 192, 
                    190, 189, 197, 187, 186, 174, 185, 195, 187, 161, 183, 173, 182, 165, 
                    185, 178, 176, 200, 187, 145, 152, 149, 149, 142, 152, 149, 152, 159, 
                    151, 148, 147, 152, 157, 158, 130, 158, 148, 146, 137, 152, 147, 143, 
                    158, 150), ncol = 4)
cov(Table37)
```

\setlength{\itemindent}{.5in}

\setlength{\itemindent}{.5in}

 Given the mean calculations above, the partitioned covariance matrix is:
 
 \[S = \begin{array}{cc|cc} 95.29 & 52.87 & 69.66 & 46.11 \\ 52.87 & 54.36 & 51.31 & 35.05 \\ \hline 69.66 & 51.31 & 100.81 & 56.54 \\ 46.11 & 35.05 & 56.54 & 45.02 \end{array} \]

